### 2021-03-08 manuscript.tex
LCNS styling, Title & Abstract

\documentclass[runningheads]{llncs}

\usepackage{graphicx}
\usepackage[hidelinks]{hyperref,xcolor}
\renewcommand\UrlFont{\color{blue}\rmfamily}

\begin{document}

\title{GeoSPARQL 1.1: an almost decadal update to the most important geospatial LOD standard}
\titlerunning{GeoSPARQL 1.1}

\author{
    Nicholas J. Car\inst{1,2}\orcidID{0000-0002-8742-7730} \and \\
    Timo Homburg\inst{3}\orcidID{1111-2222-3333-4444} \and \\
    Third Author\inst{3}\orcidID{2222-3333-4444-5555}
}

\authorrunning{Car N.J. et al.}

\institute{SURROUND Australia Pty Ltd., Brisbane Australia \and
Australian National University, Canberra Australia
\email{nicholas.car@surroundaustralia.com}\\
\url{https://surroundaustralia.com} \and
ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
\email{\{abc,lncs\}@uni-heidelberg.de}}

\maketitle

\begin{abstract}
The Open Geospatial Consortium published the GeoSPARQL 1.0 standard in 2012 containing multiple 
parts that define ``SPARQL extension functions'', ``RIF rules'', ``an RDF/OWL ontology for 
information based on the General Feature Model'' and supporting vocabularies, all for Semantic 
Web spatial data.

In the 8+ years since its publication, GeoSPARQL has become the most important spatial Semantic 
Web standard, as judged by references to it in other Semantic Web standards and its wide use in 
Semantic Web data.

An update to the standard was proposed in 2019 to deliver GeoSPARQL 1.1 in 2021 with a charter to: 
handle outstanding change requests and source new ones from the GeoSPARQL user community as well 
to ``better present'' the standard, that is to better link all the standard’s parts and better 
document \& exemplify elements. Expected updates included possible alignments to other ontologies, 
possible handling of new spatial referencing systems, new geometry representations and new artifact 
presentation.

In this paper, we will discuss the submitted change requests and resulting updates to the standard. 
We will also discuss the theory behind updates and our expectations for GeoSPARQL 1.1's use.

\keywords{GeoSPARQL  \and GeoSPARQL 1.1 \and spatial \and geospatial \and Sematnic Web \and RDF \and OWL.}
\end{abstract}

\section{First Section}
\subsection{A Subsection Sample}
Please note that the first paragraph of a section or subsection is
not indented. The first paragraph that follows a table, figure,
equation etc. does not need an indent, either.

Subsequent paragraphs, however, are indented.

\subsubsection{Sample Heading (Third Level)} Only two levels of
headings should be numbered. Lower level headings remain unnumbered;
they are formatted as run-in headings.

\paragraph{Sample Heading (Fourth Level)}
The contribution should contain no more than four levels of
headings. Table~\ref{tab1} gives a summary of all heading levels.

\begin{table}
\caption{Table captions should be placed above the
tables.}\label{tab1}
\begin{tabular}{|l|l|l|}
\hline
Heading level &  Example & Font size and style\\
\hline
Title (centered) &  {\Large\bfseries Lecture Notes} & 14 point, bold\\
1st-level heading &  {\large\bfseries 1 Introduction} & 12 point, bold\\
2nd-level heading & {\bfseries 2.1 Printing Area} & 10 point, bold\\
3rd-level heading & {\bfseries Run-in Heading in Bold.} Text follows & 10 point, bold\\
4th-level heading & {\itshape Lowest Level Heading.} Text follows & 10 point, italic\\
\hline
\end{tabular}
\end{table}

\noindent Displayed equations are centered and set on a separate
line.
\begin{equation}
x + y = z
\end{equation}
Please try to avoid rasterized images for line-art diagrams and
schemas. Whenever possible, use vector graphics instead (see
Fig.~\ref{fig1}).

\begin{figure}
\includegraphics[width=\textwidth]{fig1.eps}
\caption{A figure caption is always placed below the illustration.
Please note that short captions are centered, while long ones are
justified by the macro package automatically.} \label{fig1}
\end{figure}

\begin{theorem}
This is a sample theorem. The run-in heading is set in bold, while
the following text appears in italics. Definitions, lemmas,
propositions, and corollaries are styled the same way.
\end{theorem}
%
% the environments 'definition', 'lemma', 'proposition', 'corollary',
% 'remark', and 'example' are defined in the LLNCS documentclass as well.
%
\begin{proof}
Proofs, examples, and remarks have the initial word in italics,
while the following text appears in normal font.
\end{proof}
For citations of references, we prefer the use of square brackets
and consecutive numbers. Citations using labels or the author/year
convention are also acceptable. The following bibliography provides
a sample reference list with entries for journal
articles~\cite{ref_article1}, an LNCS chapter~\cite{ref_lncs1}, a
book~\cite{ref_book1}, proceedings without editors~\cite{ref_proc1},
and a homepage~\cite{ref_url1}. Multiple citations are grouped
\cite{ref_article1,ref_lncs1,ref_book1},
\cite{ref_article1,ref_book1,ref_proc1,ref_url1}.
%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
% \bibliographystyle{splncs04}
% \bibliography{mybibliography}
%
\begin{thebibliography}{8}
\bibitem{ref_article1}
Author, F.: Article title. Journal \textbf{2}(5), 99--110 (2016)

\bibitem{ref_lncs1}
Author, F., Author, S.: Title of a proceedings paper. In: Editor,
F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol. 9999, pp. 1--13.
Springer, Heidelberg (2016). \doi{10.10007/1234567890}

\bibitem{ref_book1}
Author, F., Author, S., Author, T.: Book title. 2nd edn. Publisher,
Location (1999)

\bibitem{ref_proc1}
Author, A.-B.: Contribution title. In: 9th International Proceedings
on Proceedings, pp. 1--2. Publisher, Location (2010)

\bibitem{ref_url1}
LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
Oct 2017
\end{thebibliography}
\end{document}

### 2021-03-08 manuscript.tex
kw tweaks

@@
 information based on the General Feature Model'' and supporting vocabularies, all for Semantic 
-Web spatial data.
+Web spatial data.\\
 
@@
 Web standard, as judged by references to it in other Semantic Web standards and its wide use in 
-Semantic Web data.
+Semantic Web data.\\
 
@@
 possible handling of new spatial referencing systems, new geometry representations and new artifact 
-presentation.
+presentation.\\
 
@@
 
-\keywords{GeoSPARQL  \and GeoSPARQL 1.1 \and spatial \and geospatial \and Sematnic Web \and RDF \and OWL.}
+\keywords{GeoSPARQL  \and GeoSPARQL 1.1 \and spatial \and geospatial \and Semantic Web \and RDF \and OWL \and OGC \and Open Geospatial Consortuim \and standard.}
 \end{abstract}

### 2021-03-08 manuscript.tex
use bib file GeoSPARQL.bib

@@
 a sample reference list with entries for journal
-articles~\cite{ref_article1}, an LNCS chapter~\cite{ref_lncs1}, a
-book~\cite{ref_book1}, proceedings without editors~\cite{ref_proc1},
-and a homepage~\cite{ref_url1}. Multiple citations are grouped
-\cite{ref_article1,ref_lncs1,ref_book1},
-\cite{ref_article1,ref_book1,ref_proc1,ref_url1}.
+articles~\cite{gonzalez-beltran_data_2018}, an LNCS chapter~\cite{gregory_todd_williams_sparql_2013}, a
+book~\cite{ref_book1}, proceedings without editors~\cite{knublauch_shapes_2017},
+and a homepage~\cite{knublauch_shapes_2017}. Multiple citations are grouped
+\cite{gonzalez-beltran_data_2018,gregory_todd_williams_sparql_2013,knublauch_shapes_2017}.
 %
@@
 %
-% \bibliographystyle{splncs04}
-% \bibliography{mybibliography}
+\bibliographystyle{splncs04}
+\bibliography{GeoSPARQL}
 %
-\begin{thebibliography}{8}
-\bibitem{ref_article1}
-Author, F.: Article title. Journal \textbf{2}(5), 99--110 (2016)
+% \begin{thebibliography}{8}
+% \bibitem{ref_article1}
+% Author, F.: Article title. Journal \textbf{2}(5), 99--110 (2016)
 
-\bibitem{ref_lncs1}
-Author, F., Author, S.: Title of a proceedings paper. In: Editor,
-F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol. 9999, pp. 1--13.
-Springer, Heidelberg (2016). \doi{10.10007/1234567890}
+% \bibitem{ref_lncs1}
+% Author, F., Author, S.: Title of a proceedings paper. In: Editor,
+% F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol. 9999, pp. 1--13.
+% Springer, Heidelberg (2016). \doi{10.10007/1234567890}
 
-\bibitem{ref_book1}
-Author, F., Author, S., Author, T.: Book title. 2nd edn. Publisher,
-Location (1999)
+% \bibitem{ref_book1}
+% Author, F., Author, S., Author, T.: Book title. 2nd edn. Publisher,
+% Location (1999)
 
-\bibitem{ref_proc1}
-Author, A.-B.: Contribution title. In: 9th International Proceedings
-on Proceedings, pp. 1--2. Publisher, Location (2010)
+% \bibitem{ref_proc1}
+% Author, A.-B.: Contribution title. In: 9th International Proceedings
+% on Proceedings, pp. 1--2. Publisher, Location (2010)
 
-\bibitem{ref_url1}
-LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
-Oct 2017
-\end{thebibliography}
+% \bibitem{ref_url1}
+% LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
+% Oct 2017
+% \end{thebibliography}
 \end{document}

### 2021-03-08 manuscript.tex
Update on Overleaf.

@@
     Nicholas J. Car\inst{1,2}\orcidID{0000-0002-8742-7730} \and \\
-    Timo Homburg\inst{3}\orcidID{1111-2222-3333-4444} \and \\
+    Timo Homburg\inst{3}\orcidID{0000-0002-9499-5840} \and \\
     Third Author\inst{3}\orcidID{2222-3333-4444-5555}
@@
 \institute{SURROUND Australia Pty Ltd., Brisbane Australia \and
-Australian National University, Canberra Australia
-\email{nicholas.car@surroundaustralia.com}\\
+Australian National University, Canberra Australia \and Mainz University Of Applied Sciences, Germany
+\email{nicholas.car@surroundaustralia.com, timo.homburg@hs-mainz.de}\\
 \url{https://surroundaustralia.com} \and

### 2021-03-10 manuscript.tex
outline comments
Update on Overleaf.

@@
+% Outline
+% 
+% Section 1
+%   History of GeoSPARWL 1.0
+% Section 2
+%   motivation to update the spec
+% Section 3
+%   GeoSPARQL 1.1 additions
+%       motivation, and expected use, of new spatial functions. In particular, why in GeoSPARQL when aggregations etc. can be performed elsewhere?
+%       Spatial Measure
+% Section 4
+%       new profile preentation
+%       new URI regimes etc
+% Section 5
+%       expected changed use modes
 \documentclass[runningheads]{llncs}
@@
 \usepackage{graphicx}
+\usepackage{todonotes}
+\usepackage{listingsutf8}
+\usepackage{cleveref}
 \usepackage[hidelinks]{hyperref,xcolor}
@@
     Nicholas J. Car\inst{1,2}\orcidID{0000-0002-8742-7730} \and \\
-    Timo Homburg\inst{3}\orcidID{0000-0002-9499-5840} \and \\
-    Third Author\inst{3}\orcidID{2222-3333-4444-5555}
+    Timo Homburg\inst{3}\orcidID{0000-0002-9499-5840} % \and \\
+    % Third Author\inst{3}\orcidID{2222-3333-4444-5555}
 }
@@
 \begin{abstract}
-The Open Geospatial Consortium published the GeoSPARQL 1.0 standard in 2012 containing multiple 
-parts that define ``SPARQL extension functions'', ``RIF rules'', ``an RDF/OWL ontology for 
+The Open Geospatial Consortium published the GeoSPARQL 1.0 standard \cite{open2012ogc} in 2012 containing multiple 
+parts that define ``SPARQL extension functions'', ``RIF rules''\cite{kifer2013rif}, ``an RDF/OWL ontology for 
 information based on the General Feature Model'' and supporting vocabularies, all for Semantic 
@@
 
-An update to the standard was proposed in 2019 to deliver GeoSPARQL 1.1 in 2021 with a charter to: 
+An update to the standard was proposed in 2019 to deliver GeoSPARQL 1.1 in 2021 with a charter\cite{abhayaratna2020ogc} to: 
 handle outstanding change requests and source new ones from the GeoSPARQL user community as well 
@@
 document \& exemplify elements. Expected updates included possible alignments to other ontologies, 
-possible handling of new spatial referencing systems, new geometry representations and new artifact 
+possible handling of new spatial referencing systems, new geometry representations, and new artifact 
 presentation.\\
@@
 \end{abstract}
+
+\section{Introduction}
+\label{sec:introduction}
+\section{Motivation to Update GeoSPARQL}
+\label{sec:motivation}
+Spatial Data On The Web Best Practices \cite{van2019best}
+GeoSPARQL White Paper \cite{geosparqlwhitepaper}
+\href{https://www.w3.org/2015/spatial/wiki/Further_development_of_GeoSPARQL}{https://www.w3.org/2015/spatial/wiki/Further\_development\_of\_GeoSPARQL}
+\section{New features in GeoSPARQL 1.1}
+\label{sec:newfeatures}
+GeoSPARQL 1.1 includes non-breaking changes to the GeoSPARQL 1.0 standard.
+\subsection{New geometry literals}
+\label{sec:literals}
+GeoSPARQL 1.1 introduces three new literal types: A GeoJSON\cite{butler2016geojson} literal, a KML\cite{nolan2014keyhole} literal, and a DGGS\cite{sahr1998discrete} WKT Literal.
+\small
+\begin{lstlisting}[caption=GeoJSON literal example,label=lst:geojsonliteral,language=sql,frame=single,basicstyle=\ttfamily]
+"{\"type\":\"Point\", \"coordinates\":[-83.38,33.95]}"
+^^<http://www.opengis.net/ont/geosparql#geoJSONLiteral>
+\end{lstlisting}
+\normalsize
+Since the GeoJSON and KML formats are restricted to be represented in the WGS84 coordinate system only, these restrictions also apply to GeoJSON and KML literals. The DGGS WKT literal  
+\subsection{Spatial aggregate functions}
+\label{sec:spatialaggregate}
+GeoSPARQL 1.1 includes new spatial aggregate functions, which allow for the quick aggregation of different geometry types.
+While spatial aggregate functions are the norm in many non-semantic geospatial databases such as POSTGIS or Oracle Spatial, at the time of defining the GeoSPARQL 1.0 standard, aggregate functions had not yet been introduced into the SPARQL standard since SPARQL 1.1 \cite{w3c_sparql_working_group_sparql_2013} was released about one year later.
+Spatial aggregate functions similar to traditional aggregate functions such as AVG,MAX, or MIN allow to aggregate results of geometry queries, e.g., to create the union out of a set of given geometry literal results.
+While calculating these aggregates may also be possible outside of the semantic database, the inclusion of the functions provides distinct advantages:
+\begin{enumerate}
+    \item No client-side library is needed to create an aggregated geometry result
+    \item Fewer and more appropriate results can be returned using a spatial aggregate function (e.g., a single union geometry vs. a set of geometries for which to create union externally)
+    \item Spatial aggregates from different SPARQL endpoints can easily be calculated using federated queries 
+\end{enumerate}
+GeoSPARQL 1.1 defines the aggregate functions \emph{geosf:boundingCircle}, calculating a bounding circle around a set of geometry, \emph{geosf:centroid} calculating the centroid of the set of geometries, \emph{geosf:concatLines} concatenating a set of linestrings that overlap, \emph{geosf:concaveHull} calculating the concave hull of a set of geometries and\emph{geosf:union}, calculating the union of a set of geometreis, in addition to the \emph{geosf:envelope} and \emph{geosf:convexHull} functions already defined in the GeoSPARQL 1.0 standard.
+Since SPARQL 1.1 these fucntions were also usable as aggregate functions.
+Spatial aggregate function definitions are accompanied by the functions \href{http://www.opengis.net/def/function/geosparql/maxX}{\emph{geof:maxX}}, \href{http://www.opengis.net/def/function/geosparql/maxY}{\emph{geof:maxY}}, \href{http://www.opengis.net/def/function/geosparql/maxZ}{\emph{geof:maxZ}} and \href{http://www.opengis.net/def/function/geosparql/minX}{\emph{geof:minX}}, \href{http://www.opengis.net/def/function/geosparql/minY}{\emph{geof:minY}}, \href{http://www.opengis.net/def/function/geosparql/minZ}{\emph{geof:minZ}} which allow to retrieve the minimum and maximum coordinates of a geometry respectively.
+\subsection{Extensions of the GeoSPARQL ontology}
+\label{sec:ontologyextensions}
+GeoSPARQL 1.1 extends the GeoSPARQL ontology by adding a new class \href{http://www.opengis.net/ont/geosparql#SpatialMeasure}{geo:SpatialMeasure}.
+This class represents a measurement such as a volume, length, or area associated with a measurement amount and a measurement unit.
+It acts as a range of three newly defined properties \href{http://www.opengis.net/ont/geosparql#hasVolume}{geo:hasArea}, \href{http://www.opengis.net/ont/geosparql#hasLength}{geo:hasLength},
+\href{http://www.opengis.net/ont/geosparql#hasVolume}{geo:hasVolume} which make these attributes of a geometry better accessible using SPARQL.
+Finally, GeoSPARQL 1.1 adds the \href{http://www.opengis.net/ont/geosparql#inCRS}{geo:inCRS} property, which allows querying the coordinate reference system of a given geometry without the need to analyze geometry literal serializations.
+The property allows for the definition of a CRS as a URI.
+It paves the way for a future definition of coordinate reference systems fully in RDF as anticipated for GeoSPARQL 2.0.
+\todo[inline]{I created a graphic (geold\_ontology.graphml) containing geo:SpatialObject, geo:Feature, geo:SpatialMeasure using the yEd Editor (https://www.yworks.com/products/yed) .
+If you find it useful we can extend it.
+}
+\begin{figure}[htb]
+    \centering
+    \includegraphics[width=\linewidth]{images/geold_ontology.png}
+    \caption{GeoSPARQL 1.1 ontology including one example feature}
+    \label{fig:geosparql11ontology}
+\end{figure}
+\section{Modernizing the documentation of GeoSPARQL}
+\label{sec:documentation}
+\subsection{Profile}
+\label{sec:profile}
+\section{Conclusions}
+\label{sec:conclusions}
+\subsection{Future Work}
+\label{sec:futurework}
+GeoSPARQL 1.2, GeoSPARQL 2.0?
+\begin{itemize}
+    \item GeoSPARQL extension ontologies: CRS systems
+    \item GeoSPARQL 1.2: More literals?
+    \item GeoSPARQL 2.0: Full featured support for 3D, simple features functions, coverages?
+\end{itemize}
+%   History of GeoSPARWL 1.0
+% Section 2
+%   motivation to update the spec
+% Section 3
+%   GeoSPARQL 1.1 additions
+%       motivation, and expected use, of new spatial functions. In particular, why in GeoSPARQL when aggregations etc. can be performed elsewhere?
+%       Spatial Measure
+% Section 4
+%       new profile preentation
+%       new URI regimes etc
+% Section 5
+%       expected changed use modes
 

### 2021-03-10 manuscript.tex
fix authors & package clashes

@@
 \usepackage{listingsutf8}
+\usepackage[hidelinks]{hyperref}
 \usepackage{cleveref}
-\usepackage[hidelinks]{hyperref,xcolor}
+
 \renewcommand\UrlFont{\color{blue}\rmfamily}
@@
 \author{
-    Nicholas J. Car\inst{1,2}\orcidID{0000-0002-8742-7730} \and \\
-    Timo Homburg\inst{3}\orcidID{0000-0002-9499-5840} % \and \\
-    % Third Author\inst{3}\orcidID{2222-3333-4444-5555}
+    Nicholas J. Car\inst{1}\orcidID{0000-0002-8742-7730} \and \\
+    Timo Homburg\inst{2}\orcidID{0000-0002-9499-5840} \and \\
+    Simon J.D Cox\inst{3}\orcidID{0000-0002-3884-3420}
 }
@@
 
-\institute{SURROUND Australia Pty Ltd., Brisbane Australia \and
-Australian National University, Canberra Australia \and Mainz University Of Applied Sciences, Germany
-\email{nicholas.car@surroundaustralia.com, timo.homburg@hs-mainz.de}\\
-\url{https://surroundaustralia.com} \and
-ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
-\email{\{abc,lncs\}@uni-heidelberg.de}}
+\institute{
+    {
+    SURROUND Australia Pty Ltd., Australia \&\\
+    Australian National University, Australia\\
+    \email{nicholas.car@surroundaustralia.com}\\
+    \url{https://surroundaustralia.com}
+    }
+    \and
+    {
+        Mainz University Of Applied Sciences, Germany\\
+        \email{timo.homburg@hs-mainz.de}
+    }
+    \and 
+    {
+        Commonwealth Scientific \& Industrial Research Organisation, Australia\\
+        \email{simon.cox@csiro.au}
+    }
+}
 

### 2021-03-11 manuscript.tex
Intro prose, some Motivation, small fixes

@@
 
-\section{Introduction}
-\label{sec:introduction}
-\section{Motivation to Update GeoSPARQL}
-\label{sec:motivation}
-Spatial Data On The Web Best Practices \cite{van2019best}
-GeoSPARQL White Paper \cite{geosparqlwhitepaper}
-\href{https://www.w3.org/2015/spatial/wiki/Further_development_of_GeoSPARQL}{https://www.w3.org/2015/spatial/wiki/Further\_development\_of\_GeoSPARQL}
-\section{New features in GeoSPARQL 1.1}
-\label{sec:newfeatures}
-GeoSPARQL 1.1 includes non-breaking changes to the GeoSPARQL 1.0 standard.
-\subsection{New geometry literals}
-\label{sec:literals}
+\section{Introduction}\label{sec:introduction}
+The GeoSPARQL standard, first issues in 2021 by the Open Geospatial Consortium (OGC)\footnote{\url{https://www.ogc.org}} 
+is one of, if not the most\footnote{It is hard to calculate use but references to GeoSPARQL in papers and other 
+well-known standards, such as DCAT2 (\url{https://www.w3.org/TR/vocab-dcat/}) suggests this}, used 
+\textit{Sematic Web} ontologies for representing spatial data.
+
+The original GeoSPARQL release, which we refere to as GeoSPARQL 1.0, contained a \textit{specification} document,
+a main ``GeoSPARQL'' ontology in an RDF file and a ``Simple Features Vocabulary'' ontology also in an RDF file. The 
+``GeoSPARQL'' ontology content, as well as lists of geospatial functions that could be performed on RDF data via 
+SPARQL\footnote{\url{https://www.w3.org/TR/sparql11-query/}} queries was defined in the specification document, as
+were entailement rules and requirements \& abstract tests for testing ontology data and function implementations. 
+Within the last few years, the function lists from the sepecification were extracted into a SKOS\footnote{\url{https://www.w3.org/TR/skos-reference/}}
+vocabulary.
+
+\section{Motivation to update GeoSPARQL}\label{sec:motivation}
+Interest in updating GeoSPARQL had been captured\footnote{\href{https://www.w3.org/2015/spatial/wiki/Further_development_of_GeoSPARQL}{https://www.w3.org/2015/spatial/wiki/Further\_development\_of\_GeoSPARQL}}
+ by the World Wide Web Consortium's (W3C) \textit{Spatial Data On The Web Working Group} (SDWWG)
+when a large body of work was done on \textit{Semantic Web} spatial data around in approximately 2015 - 2017, but 
+no updates to GeoSPARQL were ultimately made by that Working Group.
+
+% Spatial Data On The Web Best Practices \cite{van2019best}
+
+Recently, 2019, the OGC reconstituted a \textit{GeoSPARQL Standards Working Group} (SWG) to update GeoSPARQL. The general 
+motivation for work within the area of GeoSPARQL, that of \textit{Semantic Web} spatial data, and a series of
+fault fixes and proposed extensions to GeoSPARQL 1.0 are captured in an OGC White Paper \cite{geosparqlwhitepaper}. Some,
+but not all, of the SDWWG's proposals are included in the White Paper with the different communities - W3C and OGC - 
+naturally reflecting different desires.
+
+The SWG's charter - it's final scope of work - is also published by the OGC \cite{abhayaratna2020ogc} and this guides 
+the SWG's activities. Specific actions of the SWG and their staging are explained through the use of a publicly-available 
+online task tracking system within the SWG's working online code repository\footnote{}\url{https://github.com/opengeospatial/ogc-geosparql/projects/1}}.
+
+At a high-level, proposed updates to GeoSPARQL by both the SDWWG and the SWG may be categorised as:
+
+% Nick up to here
+
+\begin{itemize}
+    \item small geometry extensions, as predicted in GeoSPARQL 1.0
+    \item class extensions to cater for spatial information in RDF/OWL with more fidelity
+    \item 
+\end{itemize}
+
+\section{New features in GeoSPARQL 1.1}\label{sec:newfeatures}
+One of the first actions undertaken by the SWG was to link the GeoSPARQL 1.0 elements through a \textit{profile} 
+declaration, where a profile is a special type of \textit{specification}, as defined by \textit{The Profiles Vocabulary}\cite{atkinson_profiles_2020}. 
+The specific motivation for this was the SWG's recognition that GeoSPARQL 1.0 consisted of multiple parts, not all
+of which were easy to discover and, as a result, some GeoSPARQL users were unaware of some of the resources and some
+resources were accidentally duplicated or partly re-implemented.
+
+The profile declaration will be published by the OGC as a stand-alone resource sometime in mid-2021 along with some 
+updated GeoSPARQL 1.0 resources. Currently, all of the elements of GeoSPARQL 1.0, including the profile declaration, 
+can be found within the SWG's working online repository \footnote{\url{https://github.com/opengeospatial/ogc-geosparql}}.
+
+% GeoSPARQL 1.1 includes non-breaking changes to the GeoSPARQL 1.0 standard.
+
+\subsection{New geometry literals}\label{sec:literals}
 GeoSPARQL 1.1 introduces three new literal types: A GeoJSON\cite{butler2016geojson} literal, a KML\cite{nolan2014keyhole} literal, and a DGGS\cite{sahr1998discrete} WKT Literal.
@@
 Since the GeoJSON and KML formats are restricted to be represented in the WGS84 coordinate system only, these restrictions also apply to GeoJSON and KML literals. The DGGS WKT literal  
-\subsection{Spatial aggregate functions}
-\label{sec:spatialaggregate}
+\subsection{Spatial aggregate functions}\label{sec:spatialaggregate}
 GeoSPARQL 1.1 includes new spatial aggregate functions, which allow for the quick aggregation of different geometry types.
@@
 Spatial aggregate function definitions are accompanied by the functions \href{http://www.opengis.net/def/function/geosparql/maxX}{\emph{geof:maxX}}, \href{http://www.opengis.net/def/function/geosparql/maxY}{\emph{geof:maxY}}, \href{http://www.opengis.net/def/function/geosparql/maxZ}{\emph{geof:maxZ}} and \href{http://www.opengis.net/def/function/geosparql/minX}{\emph{geof:minX}}, \href{http://www.opengis.net/def/function/geosparql/minY}{\emph{geof:minY}}, \href{http://www.opengis.net/def/function/geosparql/minZ}{\emph{geof:minZ}} which allow to retrieve the minimum and maximum coordinates of a geometry respectively.
-\subsection{Extensions of the GeoSPARQL ontology}
-\label{sec:ontologyextensions}
+
+\subsection{Extensions of the GeoSPARQL ontology}\label{sec:ontologyextensions}
 GeoSPARQL 1.1 extends the GeoSPARQL ontology by adding a new class \href{http://www.opengis.net/ont/geosparql#SpatialMeasure}{geo:SpatialMeasure}.
@@
 \end{figure}
-\section{Modernizing the documentation of GeoSPARQL}
-\label{sec:documentation}
-\subsection{Profile}
-\label{sec:profile}
-\section{Conclusions}
-\label{sec:conclusions}
-\subsection{Future Work}
-\label{sec:futurework}
+
+\section{Modernizing the documentation of GeoSPARQL}\label{sec:documentation}
+
+\subsection{Profile}\label{sec:profile}
+
+\section{Conclusions}\label{sec:conclusions}
+
+\subsection{Future Work}\label{sec:futurework}
 GeoSPARQL 1.2, GeoSPARQL 2.0?
@@
 
-\section{First Section}
-\subsection{A Subsection Sample}
-Please note that the first paragraph of a section or subsection is
-not indented. The first paragraph that follows a table, figure,
-equation etc. does not need an indent, either.
-
-Subsequent paragraphs, however, are indented.
-
-\subsubsection{Sample Heading (Third Level)} Only two levels of
-headings should be numbered. Lower level headings remain unnumbered;
-they are formatted as run-in headings.
-
-\paragraph{Sample Heading (Fourth Level)}
-The contribution should contain no more than four levels of
-headings. Table~\ref{tab1} gives a summary of all heading levels.
-
-\begin{table}
-\caption{Table captions should be placed above the
-tables.}\label{tab1}
-\begin{tabular}{|l|l|l|}
-\hline
-Heading level &  Example & Font size and style\\
-\hline
-Title (centered) &  {\Large\bfseries Lecture Notes} & 14 point, bold\\
-1st-level heading &  {\large\bfseries 1 Introduction} & 12 point, bold\\
-2nd-level heading & {\bfseries 2.1 Printing Area} & 10 point, bold\\
-3rd-level heading & {\bfseries Run-in Heading in Bold.} Text follows & 10 point, bold\\
-4th-level heading & {\itshape Lowest Level Heading.} Text follows & 10 point, italic\\
-\hline
-\end{tabular}
-\end{table}
-
-\noindent Displayed equations are centered and set on a separate
-line.
-\begin{equation}
-x + y = z
-\end{equation}
-Please try to avoid rasterized images for line-art diagrams and
-schemas. Whenever possible, use vector graphics instead (see
-Fig.~\ref{fig1}).
-
-\begin{figure}
-\includegraphics[width=\textwidth]{fig1.eps}
-\caption{A figure caption is always placed below the illustration.
-Please note that short captions are centered, while long ones are
-justified by the macro package automatically.} \label{fig1}
-\end{figure}
-
-\begin{theorem}
-This is a sample theorem. The run-in heading is set in bold, while
-the following text appears in italics. Definitions, lemmas,
-propositions, and corollaries are styled the same way.
-\end{theorem}
-%
-% the environments 'definition', 'lemma', 'proposition', 'corollary',
-% 'remark', and 'example' are defined in the LLNCS documentclass as well.
-%
-\begin{proof}
-Proofs, examples, and remarks have the initial word in italics,
-while the following text appears in normal font.
-\end{proof}
-For citations of references, we prefer the use of square brackets
-and consecutive numbers. Citations using labels or the author/year
-convention are also acceptable. The following bibliography provides
-a sample reference list with entries for journal
-articles~\cite{gonzalez-beltran_data_2018}, an LNCS chapter~\cite{gregory_todd_williams_sparql_2013}, a
-book~\cite{ref_book1}, proceedings without editors~\cite{knublauch_shapes_2017},
-and a homepage~\cite{knublauch_shapes_2017}. Multiple citations are grouped
-\cite{gonzalez-beltran_data_2018,gregory_todd_williams_sparql_2013,knublauch_shapes_2017}.
 %

### 2021-03-11 manuscript.tex
remove citations from abstract

@@
 \begin{abstract}
-The Open Geospatial Consortium published the GeoSPARQL 1.0 standard \cite{open2012ogc} in 2012 containing multiple 
-parts that define ``SPARQL extension functions'', ``RIF rules''\cite{kifer2013rif}, ``an RDF/OWL ontology for 
+The Open Geospatial Consortium published the GeoSPARQL 1.0 standard in 2012 containing multiple 
+parts that define ``SPARQL extension functions'', ``RIF rules'', ``an RDF/OWL ontology for 
 information based on the General Feature Model'' and supporting vocabularies, all for Semantic 
@@
 
-An update to the standard was proposed in 2019 to deliver GeoSPARQL 1.1 in 2021 with a charter\cite{abhayaratna2020ogc} to: 
+An update to the standard was proposed in 2019 to deliver GeoSPARQL 1.1 in 2021 with a charter to: 
 handle outstanding change requests and source new ones from the GeoSPARQL user community as well 
@@
 the SWG's activities. Specific actions of the SWG and their staging are explained through the use of a publicly-available 
-online task tracking system within the SWG's working online code repository\footnote{}\url{https://github.com/opengeospatial/ogc-geosparql/projects/1}}.
+online task tracking system within the SWG's working online code repository\footnote{\url{https://github.com/opengeospatial/ogc-geosparql/projects/1}}.
 

### 2021-03-11 manuscript.tex
final Motivation content, linebreaking all text for easier versioning

@@
-% Outline
-% 
-% Section 1
-%   History of GeoSPARWL 1.0
-% Section 2
-%   motivation to update the spec
-% Section 3
-%   GeoSPARQL 1.1 additions
-%       motivation, and expected use, of new spatial functions. In particular, why in GeoSPARQL when aggregations etc. can be performed elsewhere?
-%       Spatial Measure
-% Section 4
-%       new profile preentation
-%       new URI regimes etc
-% Section 5
-%       expected changed use modes
 \documentclass[runningheads]{llncs}
@@
 \section{Motivation to update GeoSPARQL}\label{sec:motivation}
-Interest in updating GeoSPARQL had been captured\footnote{\href{https://www.w3.org/2015/spatial/wiki/Further_development_of_GeoSPARQL}{https://www.w3.org/2015/spatial/wiki/Further\_development\_of\_GeoSPARQL}}
+Interest in updating GeoSPARQL had been captured\footnote{\url{https://www.w3.org/2015/spatial/wiki/Further_development_of_GeoSPARQL}}
  by the World Wide Web Consortium's (W3C) \textit{Spatial Data On The Web Working Group} (SDWWG)
@@
 
-At a high-level, proposed updates to GeoSPARQL by both the SDWWG and the SWG may be categorised as:
-
-% Nick up to here
+At a high-level, proposed updates to GeoSPARQL by both the SDWWG and the SWG may be categorised as one of the following:
 
 \begin{itemize}
-    \item small geometry extensions, as predicted in GeoSPARQL 1.0
-    \item class extensions to cater for spatial information in RDF/OWL with more fidelity
-    \item 
+    \item[$\ast$] new geometry serializations
+    \begin{itemize}
+        \item[$-$] GeoJSON and other formats that have become popular since GeoSPARQL 1.0 's publication
     \end{itemize} 
+    \item[$\ast$] new ontology classes to cater for more nuanced spatial information
+    \item[$\ast$] more spatial functions
+    \begin{itemize}
+        \item[$-$] implementing functions well-known in non \texttt{Semantic Web} spatial systems
+    \end{itemize} 
+    \item[$\ast$] scalar spatial properties 
+    \begin{itemize}
+        \item[$-$] area, volume etc. alongside geometries
+    \end{itemize} 
+    \item[$\ast$] better handling of Spatial (Coordinate) Reference Systems (SRS)
+    \begin{itemize}
+        \item[$-$] potentially allowing for automated coordinate serialization conversions
+    \end{itemize} 
+    \item[$\ast$] Internet protocol-based selection of different geometries for features
+\end{itemize}
+
+Some of these propsed updates were predicted in GeoSPARQL 1.0 with the \textit{Future Work} section listing several of the 
+points above as expected or potential.
+
+The SWG's \textit{Charter}, anticipating that the more obvious updates such as new geometry serializations would certainly
+be implemented, listed the following areas of investigation that emerged from SWG proponent's discussions:
+
+\begin{itemize}
+    \item[$\ast$] a revision of the ``upper  ontology'' structuring of GeoSPARQL
+    \begin{itemize}
+        \item[$-$] better/differently defining how GeoSPARQL's \texttt{Feature} and other classes relate to one another and to other fundamental concepts in ontology
+    \end{itemize} 
+    \item[$\ast$] possible alignments to other ontologies, perhaps the \textit{W3C Time Ontology in OWL}\cite{simon_cox_time_2017}
+    \item[$\ast$] catering for new and very different SRSes, such as Discrete Global Grid Systems (DGGS) 
+\end{itemize}
+
+Specifically ruled out of scope was:
+
+\begin{itemize}
+    \item[$\ast$] investigation of property graphs
+\end{itemize}
+
+This exclusion was due to the recent (last several years) discussion in the OGC and elsewere about property graphs. The SWG 
+proponents felt that while property graphs might be important for future \textit{Semantic Web} spatial data systems, there was
+more than enough work scoped for initial SWG investigations already so as to exclude this are of investigationfor initial 
+GeoSPARQL revisions.
+
+After initial meetings, the SWG determined to make multiple releases of GeoSPARQL updates with different goals:
+
+\begin{itemize}
+    \item[$\ast$] GeoSPARQL 1.1: extensions that are fully compatable with GeoSPARQL 1.0
+    \item[$\ast$] GeoSPARQL 1.2: fully or mostly compatable extensions but which are larger additions to the standard's conceptual coverage
+    \item[$\ast$] GeoSPARQL 2.0: a future GeoSPARQL that might be quite different and partly incompatable with GeoSPARQL 1.0
+\end{itemize}
+
+The reason for expecting a future, incompatable, GeoSPARQL 2.0 is that early SWG attendees thought spatio-temporal relations
+and fundamental ontology elements in GeoSPARQL either could or should be remodelled which might break the current, familiar, 
+\texttt{Feature}/\texttt{Geometry} class relations. Details of these potential changes haven't been fully expounded, at the 
+time of this paper, however initial SWG attendees' intuition is that a future GeoSPARQL 2.0 might generalise spatial concepts and
+move away from only, or primarily, \textit{geo}spatial, or perhaps focus not just on \texttt{Feature}/\texttt{Geometry} relations
+but look to generalised mecahnisms for describing dimensions of features of which \textit{geometry} is one of many (and perhaps)
+\textit{temporality} too).
+
+An additional area of updates to GeoSPARQL that was not forseen by GeoSPARQL 1.0, the SDWWG or the initial proponants of the SWG
+but which was included by the early SWG attendees in GeosPARQL 1.1 and is now present in the SWG's working repository was that 
+of new modes of standard presentation. The motivation for this was conceptual work within the W3C and the OGC to do with how 
+multi-part standards might link their parts and declare dependencies. This resulted in the profile declaration work, explained 
+in the next section.
 
@@
 \subsection{New geometry literals}\label{sec:literals}
-GeoSPARQL 1.1 introduces three new literal types: A GeoJSON\cite{butler2016geojson} literal, a KML\cite{nolan2014keyhole} literal, and a DGGS\cite{sahr1998discrete} WKT Literal.
+GeoSPARQL 1.1 introduces three new literal types: A GeoJSON\cite{butler2016geojson} literal, a KML\cite{nolan2014keyhole} 
+literal, and a DGGS\cite{sahr1998discrete} WKT Literal.
+
 \small
@@
 \normalsize
-Since the GeoJSON and KML formats are restricted to be represented in the WGS84 coordinate system only, these restrictions also apply to GeoJSON and KML literals. The DGGS WKT literal  
+Since the GeoJSON and KML formats are restricted to be represented in the WGS84 SRS only, these restrictions 
+also apply to GeoJSON and KML literals. The DGGS WKT literal  
 \subsection{Spatial aggregate functions}\label{sec:spatialaggregate}
 GeoSPARQL 1.1 includes new spatial aggregate functions, which allow for the quick aggregation of different geometry types. 
-While spatial aggregate functions are the norm in many non-semantic geospatial databases such as POSTGIS or Oracle Spatial, at the time of defining the GeoSPARQL 1.0 standard, aggregate functions had not yet been introduced into the SPARQL standard since SPARQL 1.1 \cite{w3c_sparql_working_group_sparql_2013} was released about one year later.
-Spatial aggregate functions similar to traditional aggregate functions such as AVG,MAX, or MIN allow to aggregate results of geometry queries, e.g., to create the union out of a set of given geometry literal results.
-While calculating these aggregates may also be possible outside of the semantic database, the inclusion of the functions provides distinct advantages:
+While spatial aggregate functions are the norm in many non-semantic geospatial databases such as POSTGIS or Oracle Spatial, 
+at the time of defining the GeoSPARQL 1.0 standard, aggregate functions had not yet been introduced into the SPARQL standard 
+since SPARQL 1.1 \cite{w3c_sparql_working_group_sparql_2013} was released about one year later. Spatial aggregate functions 
+similar to traditional aggregate functions such as AVG,MAX, or MIN allow to aggregate results of geometry queries, e.g., to 
+create the union out of a set of given geometry literal results. While calculating these aggregates may also be possible 
+outside of the semantic database, the inclusion of the functions provides distinct advantages:
 \begin{enumerate}
@@
 \end{enumerate}
-GeoSPARQL 1.1 defines the aggregate functions \emph{geosf:boundingCircle}, calculating a bounding circle around a set of geometry, \emph{geosf:centroid} calculating the centroid of the set of geometries, \emph{geosf:concatLines} concatenating a set of linestrings that overlap, \emph{geosf:concaveHull} calculating the concave hull of a set of geometries and\emph{geosf:union}, calculating the union of a set of geometreis, in addition to the \emph{geosf:envelope} and \emph{geosf:convexHull} functions already defined in the GeoSPARQL 1.0 standard.
-Since SPARQL 1.1 these fucntions were also usable as aggregate functions.
-Spatial aggregate function definitions are accompanied by the functions \href{http://www.opengis.net/def/function/geosparql/maxX}{\emph{geof:maxX}}, \href{http://www.opengis.net/def/function/geosparql/maxY}{\emph{geof:maxY}}, \href{http://www.opengis.net/def/function/geosparql/maxZ}{\emph{geof:maxZ}} and \href{http://www.opengis.net/def/function/geosparql/minX}{\emph{geof:minX}}, \href{http://www.opengis.net/def/function/geosparql/minY}{\emph{geof:minY}}, \href{http://www.opengis.net/def/function/geosparql/minZ}{\emph{geof:minZ}} which allow to retrieve the minimum and maximum coordinates of a geometry respectively.
+GeoSPARQL 1.1 defines the aggregate functions \emph{geosf:boundingCircle}, calculating a bounding circle around a set of 
+geometry, \emph{geosf:centroid} calculating the centroid of the set of geometries, \emph{geosf:concatLines} concatenating 
+a set of linestrings that overlap, \emph{geosf:concaveHull} calculating the concave hull of a set of geometries and\emph{geosf:union}, 
+calculating the union of a set of geometreis, in addition to the \emph{geosf:envelope} and \emph{geosf:convexHull} functions 
+already defined in the GeoSPARQL 1.0 standard. Since SPARQL 1.1 these fucntions were also usable as aggregate functions.
+
+Spatial aggregate function definitions are accompanied by the functions \href{http://www.opengis.net/def/function/geosparql/maxX}{\emph{geof:maxX}}, 
+\href{http://www.opengis.net/def/function/geosparql/maxY}{\emph{geof:maxY}}, \href{http://www.opengis.net/def/function/geosparql/maxZ}{\emph{geof:maxZ}} 
+and \href{http://www.opengis.net/def/function/geosparql/minX}{\emph{geof:minX}}, \href{http://www.opengis.net/def/function/geosparql/minY}{\emph{geof:minY}}, 
+\href{http://www.opengis.net/def/function/geosparql/minZ}{\emph{geof:minZ}} which allow to retrieve the minimum and maximum 
+coordinates of a geometry respectively.
 
@@
 GeoSPARQL 1.1 extends the GeoSPARQL ontology by adding a new class \href{http://www.opengis.net/ont/geosparql#SpatialMeasure}{geo:SpatialMeasure}. 
-This class represents a measurement such as a volume, length, or area associated with a measurement amount and a measurement unit.
-It acts as a range of three newly defined properties \href{http://www.opengis.net/ont/geosparql#hasVolume}{geo:hasArea}, \href{http://www.opengis.net/ont/geosparql#hasLength}{geo:hasLength},
-\href{http://www.opengis.net/ont/geosparql#hasVolume}{geo:hasVolume} which make these attributes of a geometry better accessible using SPARQL.
-Finally, GeoSPARQL 1.1 adds the \href{http://www.opengis.net/ont/geosparql#inCRS}{geo:inCRS} property, which allows querying the coordinate reference system of a given geometry without the need to analyze geometry literal serializations.
-The property allows for the definition of a CRS as a URI.
-It paves the way for a future definition of coordinate reference systems fully in RDF as anticipated for GeoSPARQL 2.0.
-\todo[inline]{I created a graphic (geold\_ontology.graphml) containing geo:SpatialObject, geo:Feature, geo:SpatialMeasure using the yEd Editor (https://www.yworks.com/products/yed) .
-If you find it useful we can extend it.
-}
+This class represents a measurement such as a volume, length, or area associated with a measurement amount and a measurement 
+unit. It acts as a range of three newly defined properties \href{http://www.opengis.net/ont/geosparql#hasVolume}{geo:hasArea}, 
+\href{http://www.opengis.net/ont/geosparql#hasLength}{geo:hasLength}, \href{http://www.opengis.net/ont/geosparql#hasVolume}{geo:hasVolume} 
+which make these attributes of a geometry better accessible using SPARQL. Finally, GeoSPARQL 1.1 adds the 
+\href{http://www.opengis.net/ont/geosparql#inSRS}{geo:inSRS} property, which allows querying the SRS of a given geometry 
+without the need to analyze geometry literal serializations. The property allows for the definition of a SRS as a URI. It 
+paves the way for a future definition of SRS fully in RDF as anticipated for GeoSPARQL 2.0.
+
 \begin{figure}[htb]
@@
 \begin{itemize}
-    \item GeoSPARQL extension ontologies: CRS systems
+    \item GeoSPARQL extension ontologies: SRS systems
     \item GeoSPARQL 1.2: More literals?
@@
 \end{document}
+
+% Outline
+% 
+% Section 1
+%   History of GeoSPARWL 1.0
+% Section 2
+%   motivation to update the spec
+% Section 3
+%   GeoSPARQL 1.1 additions
+%       motivation, and expected use, of new spatial functions. In particular, why in GeoSPARQL when aggregations etc. can be performed elsewhere?
+%       Spatial Measure
+% Section 4
+%       new profile preentation
+%       new URI regimes etc
+% Section 5
+%       expected changed use modes

### 2021-03-11 manuscript.tex
Updates in GSP 1.1

@@
 The GeoSPARQL standard, first issues in 2021 by the Open Geospatial Consortium (OGC)\footnote{\url{https://www.ogc.org}} 
-is one of, if not the most\footnote{It is hard to calculate use but references to GeoSPARQL in papers and other 
-well-known standards, such as DCAT2 (\url{https://www.w3.org/TR/vocab-dcat/}) suggests this}, used 
-\textit{Sematic Web} ontologies for representing spatial data.
+is one of, if not the most\footnote{It's hard to calculate use but references to GeoSPARQL other well-known standards, 
+such as DCAT2 (\url{https://www.w3.org/TR/vocab-dcat/}) suggests this}, used \textit{Sematic Web} ontologies for 
+representing spatial data.
 
@@
 
-Specifically ruled out of scope was:
-
-\begin{itemize}
-    \item[$\ast$] investigation of property graphs
-\end{itemize}
-
-This exclusion was due to the recent (last several years) discussion in the OGC and elsewere about property graphs. The SWG 
-proponents felt that while property graphs might be important for future \textit{Semantic Web} spatial data systems, there was
-more than enough work scoped for initial SWG investigations already so as to exclude this are of investigationfor initial 
-GeoSPARQL revisions.
+Specifically ruled out of scope was any investigation of property graphs. Recent (last several years) discussion in the OGC and 
+elsewere about property graphs moticated a consideration of them, hoever the SWG proponents felt that while property graphs might 
+be important for future \textit{Semantic Web} spatial data systems, there was more than enough work scoped for initial SWG work
+(several revisions of the standard) to initially exclude this are of investigation.
 
@@
 \begin{itemize}
-    \item[$\ast$] GeoSPARQL 1.1: extensions that are fully compatable with GeoSPARQL 1.0
-    \item[$\ast$] GeoSPARQL 1.2: fully or mostly compatable extensions but which are larger additions to the standard's conceptual coverage
-    \item[$\ast$] GeoSPARQL 2.0: a future GeoSPARQL that might be quite different and partly incompatable with GeoSPARQL 1.0
+    \item[$\ast$] \textbf{1.1}: extensions that are fully compatable with GeoSPARQL 1.0
+    \item[$\ast$] \textbf{1.2}: fully or mostly compatable extensions but which are larger additions to the standard's conceptual coverage
+    \item[$\ast$] \textbf{2.0}: a future GeoSPARQL that might be quite different and partly incompatable with GeoSPARQL 1.0
 \end{itemize}
@@
 move away from only, or primarily, \textit{geo}spatial, or perhaps focus not just on \texttt{Feature}/\texttt{Geometry} relations
-but look to generalised mecahnisms for describing dimensions of features of which \textit{geometry} is one of many (and perhaps)
+but look to generalised mechanisms for describing dimensions of features of which \textit{geometry} is one of many (and perhaps)
 \textit{temporality} too).
@@
 
-\section{New features in GeoSPARQL 1.1}\label{sec:newfeatures}
+\section{Updates in GeoSPARQL 1.1}\label{sec:newfeatures}
+So far (as of June, 2020) the GeoSPARQL SWG has triarged changed requests into GeoSPARQL 1.1, 1.2 \$ 2.0 releases and has addressed
+many 1.1 requests. Here we report only on those 1.1 requests. This section describes some of the areas updated in GeoSAPRQL 1.1
+and the following section foreshadows likely 1.2 and 2.0 updates.
+
+\subsection{Profile Declaration}\label{sec:profiledec}
 One of the first actions undertaken by the SWG was to link the GeoSPARQL 1.0 elements through a \textit{profile} 
@@
 of which were easy to discover and, as a result, some GeoSPARQL users were unaware of some of the resources and some
-resources were accidentally duplicated or partly re-implemented.
+resources were accidentally duplicated or partly re-implemented. Profile declarations of this sort are anticipated, by the OGC, 
+as being the \textit{best practice} way for it to deliver multi-part standards.
 
-The profile declaration will be published by the OGC as a stand-alone resource sometime in mid-2021 along with some 
+The profile declaration for GeosPARQL 1.0 will be published by the OGC as a stand-alone resource sometime in early 2021 along with some 
 updated GeoSPARQL 1.0 resources. Currently, all of the elements of GeoSPARQL 1.0, including the profile declaration, 
 can be found within the SWG's working online repository \footnote{\url{https://github.com/opengeospatial/ogc-geosparql}}.
+The profile declaration for GeosPARQL 1.1 will be published at the same time as all, or most, of the 1.1 releases' updated
+resources, currently expected in mid-2021.
 
-% GeoSPARQL 1.1 includes non-breaking changes to the GeoSPARQL 1.0 standard.
+\subsection{New geometry literals}\label{sec:newliterals}
+GeoSPARQL 1.1 introduces three new litergeometry serializations:
 
-\subsection{New geometry literals}\label{sec:literals}
-GeoSPARQL 1.1 introduces three new literal types: A GeoJSON\cite{butler2016geojson} literal, a KML\cite{nolan2014keyhole} 
-literal, and a DGGS\cite{sahr1998discrete} WKT Literal.
+\begin{enumerate}
+    \item GeoJSON (Geo- JavaScript Object Notation)\cite{butler2016geojson}
+    \item KML (Keyhole Markup Language)\cite{nolan2014keyhole} 
+    \item DGGS (Discrete Global Grid System)\cite{sahr1998discrete}
+\end{enumerate}
+
+The first two of these are ``obvious'' in the sence that that they have been expected/requested by the SDWWG and many users
+of GeoSPARQL for a long time, due to those format's popularity. an example of a GeoJSON geometry serialization is given below.
 
 \small
-\begin{lstlisting}[caption=GeoJSON literal example,label=lst:geojsonliteral,language=sql,frame=single,basicstyle=\ttfamily]
-"{\"type\":\"Point\", \"coordinates\":[-83.38,33.95]}"
+\begin{lstlisting}[caption=GeoJSON geometry serialization example,label=lst:geojsonliteral,language=sql,frame=single,basicstyle=\ttfamily]
+"""{"type":"Point", "coordinates":[-83.38,33.95]}"""
 ^^<http://www.opengis.net/ont/geosparql#geoJSONLiteral>
@@
 also apply to GeoJSON and KML literals. The DGGS WKT literal
-\subsection{Spatial aggregate functions}\label{sec:spatialaggregate}
+
+The third new serialization format is more forward-looking in that it is not dirven by user demand but by predected demand.
+DGGS does not have a single, concrete format standard as the others do, not is it ever likely to - different DGGSes will 
+likely implement very different data formats - so GeoSAPRQL 1.1 makes generalized provisions for DGGS serializations but 
+presents no detailed requirements for them, only stating that the specific DGGS must be identified.
+
+\subsection{New spatial functions}\label{sec:newfunctions}
 GeoSPARQL 1.1 includes new spatial aggregate functions, which allow for the quick aggregation of different geometry types. 
@@
 outside of the semantic database, the inclusion of the functions provides distinct advantages:
+
 \begin{enumerate}
@@
 \end{enumerate}
+
 GeoSPARQL 1.1 defines the aggregate functions \emph{geosf:boundingCircle}, calculating a bounding circle around a set of 
@@
 
-\subsection{Extensions of the GeoSPARQL ontology}\label{sec:ontologyextensions}
+\subsection{Ontology extensions}\label{sec:ontexts}
 GeoSPARQL 1.1 extends the GeoSPARQL ontology by adding a new class \href{http://www.opengis.net/ont/geosparql#SpatialMeasure}{geo:SpatialMeasure}. 

### 2021-03-11 manuscript.tex
text improvements through to Conclusion

@@
 The profile declaration for GeosPARQL 1.0 will be published by the OGC as a stand-alone resource sometime in early 2021 along with some 
-updated GeoSPARQL 1.0 resources. Currently, all of the elements of GeoSPARQL 1.0, including the profile declaration, 
-can be found within the SWG's working online repository \footnote{\url{https://github.com/opengeospatial/ogc-geosparql}}.
-The profile declaration for GeosPARQL 1.1 will be published at the same time as all, or most, of the 1.1 releases' updated
-resources, currently expected in mid-2021.
+updated GeoSPARQL 1.0 resources. The profile declaration for GeosPARQL 1.1 will be published at the same time as all, or most, of the 
+1.1 releases' updated resources, currently expected in mid-2021. The 1.1 releases' resources are:
+
+\begin{enumerate}
+    \item a profile declaration
+    \item a specification document
+    \item an RDF/OWL ontology document
+    \item a Functions \& Rules vocabulary, derived from the ontology
+    \item a Simple Features feature types vocabulary
+    \item a set of RIF\cite{kifer2013rif} rules
+    \item SHACL\cite{knublauch_shapes_2017} shapes for RDF data validation 
+\end{enumerate}
+
+Both the 1.0 and the 1.1 resources are currently avaialbe from the SWG's online code repository\footnote{\url{https://github.com/opengeospatial/ogc-geosparql}}.
 
 \subsection{New geometry literals}\label{sec:newliterals}
-GeoSPARQL 1.1 introduces three new litergeometry serializations:
+Three new geometry serializations are introduced:
 
 \begin{enumerate}
-    \item GeoJSON (Geo- JavaScript Object Notation)\cite{butler2016geojson}
-    \item KML (Keyhole Markup Language)\cite{nolan2014keyhole} 
-    \item DGGS (Discrete Global Grid System)\cite{sahr1998discrete}
+    \item \textbf{GeoJSON} (Geo- JavaScript Object Notation)\cite{butler2016geojson}
+    \item \textbf{KML} (Keyhole Markup Language)\cite{nolan2014keyhole} 
+    \item \textbf{DGGS} (Discrete Global Grid System)\cite{sahr1998discrete}
 \end{enumerate}
 
-The first two of these are ``obvious'' in the sence that that they have been expected/requested by the SDWWG and many users
-of GeoSPARQL for a long time, due to those format's popularity. an example of a GeoJSON geometry serialization is given below.
+An example of a point's GeoJSON geometry serialization is given below, followed by an unrelated simple polygon \textit{AusPIX}
+\footnote{\url{https://w3id.org/dggs/auspix}} DGGS geometry serialization.
 
 \small
-\begin{lstlisting}[caption=GeoJSON geometry serialization example,label=lst:geojsonliteral,language=sql,frame=single,basicstyle=\ttfamily]
+\begin{lstlisting}[caption=GeoJSON geometry serialization example,label=lst:geojsonliteral,language=sql,frame=single,basicstyle=\ttfamily,showstringspaces=false]
 """{"type":"Point", "coordinates":[-83.38,33.95]}"""
@@
 \end{lstlisting}
+
+\begin{lstlisting}[caption=AusPIX DGGS geometry serialization example,label=lst:geodggsWktliteral,language=sql,frame=single,basicstyle=\ttfamily,showstringspaces=false]
+"""<https://w3id.org/dggs/auspix> DirectedOrdinateList 
+(R3231 R3234 R3235 R3238 R3243 R3246)"""
+^^<http://www.opengis.net/ont/geosparql#dggsWktLiteral>
+\end{lstlisting}
 \normalsize
-Since the GeoJSON and KML formats are restricted to be represented in the WGS84 SRS only, these restrictions 
-also apply to GeoJSON and KML literals. The DGGS WKT literal
+
+The first two of these new serializations have been much anticipated and were requested by the SDWWG and many users of 
+GeoSPARQL, due to those formats' popularity.
 
@@
 \subsection{New spatial functions}\label{sec:newfunctions}
-GeoSPARQL 1.1 includes new spatial aggregate functions, which allow for the quick aggregation of different geometry types. 
-While spatial aggregate functions are the norm in many non-semantic geospatial databases such as POSTGIS or Oracle Spatial, 
-at the time of defining the GeoSPARQL 1.0 standard, aggregate functions had not yet been introduced into the SPARQL standard 
-since SPARQL 1.1 \cite{w3c_sparql_working_group_sparql_2013} was released about one year later. Spatial aggregate functions 
-similar to traditional aggregate functions such as AVG,MAX, or MIN allow to aggregate results of geometry queries, e.g., to 
-create the union out of a set of given geometry literal results. While calculating these aggregates may also be possible 
-outside of the semantic database, the inclusion of the functions provides distinct advantages:
+While spatial aggregation functions are the norm in many non-semantic geospatial databases such as POSTGIS or Oracle Spatial, 
+at the time of defining the GeoSPARQL 1.0 standard, aggregation functions had not yet been introduced into the SPARQL standard,
+but have been since with SPARQL 1.1~\cite{w3c_sparql_working_group_sparql_2013}. Spatial aggregation functions 
+similar to traditional (relation database) aggregation functions such as AVG, MAX, or MIN allow aggregated results of geometry 
+queries, for example to create the union of a set of selected serialized geometries. While calculating these aggregates is 
+certainly possible outside of a semantic database, and thus GeoSPARQL, the inclusion of the functions provides distinct advantages:
 
@@
     \item No client-side library is needed to create an aggregated geometry result
-    \item Fewer and more appropriate results can be returned using a spatial aggregate function (e.g., a single union geometry vs. a set of geometries for which to create union externally)
+    \item Fewer and more appropriate results can be returned using a spatial aggregate function (for example, a union geometry result)
     \item Spatial aggregates from different SPARQL endpoints can easily be calculated using federated queries 
@@
 
-GeoSPARQL 1.1 defines the aggregate functions \emph{geosf:boundingCircle}, calculating a bounding circle around a set of 
-geometry, \emph{geosf:centroid} calculating the centroid of the set of geometries, \emph{geosf:concatLines} concatenating 
-a set of linestrings that overlap, \emph{geosf:concaveHull} calculating the concave hull of a set of geometries and\emph{geosf:union}, 
-calculating the union of a set of geometreis, in addition to the \emph{geosf:envelope} and \emph{geosf:convexHull} functions 
-already defined in the GeoSPARQL 1.0 standard. Since SPARQL 1.1 these fucntions were also usable as aggregate functions.
+In addition to the \emph{geosf:union}, \emph{geosf:envelope} and \emph{geosf:convexHull} functions already defined in GeoSPARQL 1.0, 
+for use within SPARQL \texttt{FILTER} operations, GeoSPARQL 1.1 defines \emph{geosf:Union}, \emph{geosf:Envelope} and \emph{geosf:ConvexHull}
+as well as \emph{geosf:BoundingCircle}, \emph{geosf:Centroid}, \emph{geosf:ConcatLines} - concatenating a set of overlapping linestrings 
+that overlap - and \emph{geosf:ConcaveHull} that can be used to return aggregated results. Listing~\ref{lst:agfunc} gives an example 
+of one of these new functions in use.
 
-Spatial aggregate function definitions are accompanied by the functions \href{http://www.opengis.net/def/function/geosparql/maxX}{\emph{geof:maxX}}, 
-\href{http://www.opengis.net/def/function/geosparql/maxY}{\emph{geof:maxY}}, \href{http://www.opengis.net/def/function/geosparql/maxZ}{\emph{geof:maxZ}} 
-and \href{http://www.opengis.net/def/function/geosparql/minX}{\emph{geof:minX}}, \href{http://www.opengis.net/def/function/geosparql/minY}{\emph{geof:minY}}, 
-\href{http://www.opengis.net/def/function/geosparql/minZ}{\emph{geof:minZ}} which allow to retrieve the minimum and maximum 
-coordinates of a geometry respectively.
+\small
+\begin{lstlisting}[caption=Aggregation Function example SPARQL query,label=lst:agfunc,language=sql,frame=single,basicstyle=\ttfamily,showstringspaces=false]
+# returns the circle geometry bounding all the geometries 
+# of Feature <x>
+SELECT (geosaf:BoundingCircle(?geo) AS ?circ)
+WHERE {
+    <x> geo:hasGeometry/geo:asWKT ?geo .
+}
+\end{lstlisting}
+\normalsize
+
+New functions to retieve the minimum and maximum values of geometries' coordinates are also added: \emph{geof:minX} \& \emph{geof:maxX},
+\emph{geof:minY} \& \emph{geof:maxY} and \emph{geof:minZ} \& \emph{geof:maxZ}.
 
 \subsection{Ontology extensions}\label{sec:ontexts}
-GeoSPARQL 1.1 extends the GeoSPARQL ontology by adding a new class \href{http://www.opengis.net/ont/geosparql#SpatialMeasure}{geo:SpatialMeasure}. 
-This class represents a measurement such as a volume, length, or area associated with a measurement amount and a measurement 
-unit. It acts as a range of three newly defined properties \href{http://www.opengis.net/ont/geosparql#hasVolume}{geo:hasArea}, 
-\href{http://www.opengis.net/ont/geosparql#hasLength}{geo:hasLength}, \href{http://www.opengis.net/ont/geosparql#hasVolume}{geo:hasVolume} 
-which make these attributes of a geometry better accessible using SPARQL. Finally, GeoSPARQL 1.1 adds the 
-\href{http://www.opengis.net/ont/geosparql#inSRS}{geo:inSRS} property, which allows querying the SRS of a given geometry 
-without the need to analyze geometry literal serializations. The property allows for the definition of a SRS as a URI. It 
-paves the way for a future definition of SRS fully in RDF as anticipated for GeoSPARQL 2.0.
+GeoSPARQL 1.1 extends the GeoSPARQL ontology by adding a new class, \texttt{geo:SpatialMeasure}. This class represents a spatial measurement 
+such as a volume, length, or area associated with a measurement amount and a unit of measure. It is the range of three newly-defined properties:
+\texttt{geo:hasArea}, \texttt{geo:hasLength} and \texttt{geo:hasVolume} which make these attributes of a geometry better accessible using 
+SPARQL.
+
+These additions address a number of change requests from the SDWWG and the SWG but, given the introduction of a dedicated \textit{measure} class, 
+the do more than just provide for a few scalar spatial measures and open up GeoSPARQL to general patterns of measurement present in ontologies 
+such as the W3C's \textit{SOSA}~\cite{haller_semantic_2017}.
+
+The 1.1 release also adds the property \texttt{geo:inSRS}, which allows querying for the SRS of a given geometry without the need to analyze 
+geometry serializations. The property allows for the definition of a SRS as an Internationalized Resource Identifier (IRI). It paves the way 
+for a future definition of SRS fully in RDF, as anticipated for GeoSPARQL 2.0. Figure~\ref{fig:geosparql11ontology} overviews the ontology, as 
+of release 1.1.
 
@@
 \end{figure}
-
-\section{Modernizing the documentation of GeoSPARQL}\label{sec:documentation}
-
-\subsection{Profile}\label{sec:profile}
 

### 2021-03-11 manuscript.tex
Update on Overleaf.

@@
 
-\keywords{GeoSPARQL  \and GeoSPARQL 1.1 \and spatial \and geospatial \and Semantic Web \and RDF \and OWL \and OGC \and Open Geospatial Consortuim \and standard.}
+\keywords{GeoSPARQL  \and GeoSPARQL 1.1 \and spatial \and geospatial \and Semantic Web \and RDF \and OWL \and OGC \and Open Geospatial Consortium \and standard.}
 \end{abstract}
@@
 
-The original GeoSPARQL release, which we refere to as GeoSPARQL 1.0, contained a \textit{specification} document,
+The original GeoSPARQL release, which we refer to as GeoSPARQL 1.0, contained a \textit{specification} document,
 a main ``GeoSPARQL'' ontology in an RDF file and a ``Simple Features Vocabulary'' ontology also in an RDF file. The 
@@
 SPARQL\footnote{\url{https://www.w3.org/TR/sparql11-query/}} queries was defined in the specification document, as
-were entailement rules and requirements \& abstract tests for testing ontology data and function implementations. 
-Within the last few years, the function lists from the sepecification were extracted into a SKOS\footnote{\url{https://www.w3.org/TR/skos-reference/}}
+were entailment rules and requirements \& abstract tests for testing ontology data and function implementations. 
+Within the last few years, the function lists from the specification were extracted into a SKOS\footnote{\url{https://www.w3.org/TR/skos-reference/}}
 vocabulary.
@@
 fault fixes and proposed extensions to GeoSPARQL 1.0 are captured in an OGC White Paper \cite{geosparqlwhitepaper}. Some,
-but not all, of the SDWWG's proposals are included in the White Paper with the different communities - W3C and OGC - 
+but not all of the SDWWG's proposals are included in the White Paper with the different communities - W3C and OGC - 
 naturally reflecting different desires.
 
-The SWG's charter - it's final scope of work - is also published by the OGC \cite{abhayaratna2020ogc} and this guides 
+The SWG's charter - its final scope of work - is also published by the OGC \cite{abhayaratna2020ogc} and this guides 
 the SWG's activities. Specific actions of the SWG and their staging are explained through the use of a publicly-available 
@@
 
-Some of these propsed updates were predicted in GeoSPARQL 1.0 with the \textit{Future Work} section listing several of the 
+Some of these proposed updates were predicted in GeoSPARQL 1.0, with the \textit{Future Work} section listing several of the 
 points above as expected or potential.
@@
     \begin{itemize}
-        \item[$-$] better/differently defining how GeoSPARQL's \texttt{Feature} and other classes relate to one another and to other fundamental concepts in ontology
+        \item[$-$] better/differently defining how GeoSPARQL's \texttt{Feature} and other classes relate to one another and to other fundamental concepts in the ontology
     \end{itemize} 
@@
 Specifically ruled out of scope was any investigation of property graphs. Recent (last several years) discussion in the OGC and 
-elsewere about property graphs moticated a consideration of them, hoever the SWG proponents felt that while property graphs might 
+elsewhere about property graphs motivated a consideration of them, however, the SWG proponents felt that while property graphs might 
 be important for future \textit{Semantic Web} spatial data systems, there was more than enough work scoped for initial SWG work
-(several revisions of the standard) to initially exclude this are of investigation.
+(several revisions of the standard) to initially exclude this area of investigation.
 
@@
 
-The reason for expecting a future, incompatable, GeoSPARQL 2.0 is that early SWG attendees thought spatio-temporal relations
-and fundamental ontology elements in GeoSPARQL either could or should be remodelled which might break the current, familiar, 
+The reason for expecting a future, incompatible, GeoSPARQL 2.0 is that early SWG attendees thought spatio-temporal relations
+and fundamental ontology elements in GeoSPARQL either could or should be remodelled, which might break the current, familiar, 
 \texttt{Feature}/\texttt{Geometry} class relations. Details of these potential changes haven't been fully expounded, at the 
@@
 
-An additional area of updates to GeoSPARQL that was not forseen by GeoSPARQL 1.0, the SDWWG or the initial proponants of the SWG
-but which was included by the early SWG attendees in GeosPARQL 1.1 and is now present in the SWG's working repository was that 
+An additional area of updates to GeoSPARQL that was not foreseen by GeoSPARQL 1.0, the SDWWG, or the initial proponents of the SWG
+but which was included by the early SWG attendees in GeoSPARQL 1.1 and is now present in the SWG's working repository was that 
 of new modes of standard presentation. The motivation for this was conceptual work within the W3C and the OGC to do with how 
@@
 \section{Updates in GeoSPARQL 1.1}\label{sec:newfeatures}
-So far (as of June, 2020) the GeoSPARQL SWG has triarged changed requests into GeoSPARQL 1.1, 1.2 \$ 2.0 releases and has addressed
+So far (as of June, 2020) the GeoSPARQL SWG has triaged changed requests into GeoSPARQL 1.1, 1.2 \$ 2.0 releases and has addressed
 many 1.1 requests. Here we report only on those 1.1 requests. This section describes some of the areas updated in GeoSAPRQL 1.1
@@
 The specific motivation for this was the SWG's recognition that GeoSPARQL 1.0 consisted of multiple parts, not all
-of which were easy to discover and, as a result, some GeoSPARQL users were unaware of some of the resources and some
+of which were easy to discover. As a result, some GeoSPARQL users were unaware of some of the resources and some
 resources were accidentally duplicated or partly re-implemented. Profile declarations of this sort are anticipated, by the OGC, 
@@
 
-Both the 1.0 and the 1.1 resources are currently avaialbe from the SWG's online code repository\footnote{\url{https://github.com/opengeospatial/ogc-geosparql}}.
+Both the 1.0 and the 1.1 resources are currently available from the SWG's online code repository\footnote{\url{https://github.com/opengeospatial/ogc-geosparql}}.
 
@@
 
-The third new serialization format is more forward-looking in that it is not dirven by user demand but by predected demand.
-DGGS does not have a single, concrete format standard as the others do, not is it ever likely to - different DGGSes will 
+The third new serialization format is more forward-looking in that it is not driven by user demand but by predicted demand.
+DGGS does not have a single, concrete format standard as the others do, nor is it ever likely to - different DGGSes will 
 likely implement very different data formats - so GeoSAPRQL 1.1 makes generalized provisions for DGGS serializations but 
@@
 but have been since with SPARQL 1.1~\cite{w3c_sparql_working_group_sparql_2013}. Spatial aggregation functions 
-similar to traditional (relation database) aggregation functions such as AVG, MAX, or MIN allow aggregated results of geometry 
-queries, for example to create the union of a set of selected serialized geometries. While calculating these aggregates is 
+similar to traditional (relation database), aggregation functions such as AVG, MAX, or MIN allow aggregated results of geometry 
+queries, for example, to create the union of a set of selected serialized geometries. While calculating these aggregates is 
 certainly possible outside of a semantic database, and thus GeoSPARQL, the inclusion of the functions provides distinct advantages:
@@
 
-In addition to the \emph{geosf:union}, \emph{geosf:envelope} and \emph{geosf:convexHull} functions already defined in GeoSPARQL 1.0, 
+In addition to the \href{http://www.opengis.net/def/function/geosparql/union}{\emph{geof:union}}, \href{http://www.opengis.net/def/function/geosparql/envelope}{\emph{geof:envelope}} and \href{http://www.opengis.net/def/function/geosparql/convexHull}{\emph{geof:convexHull}} functions already defined in GeoSPARQL 1.0,
 for use within SPARQL \texttt{FILTER} operations, GeoSPARQL 1.1 defines \emph{geosf:Union}, \emph{geosf:Envelope} and \emph{geosf:ConvexHull}
@@
 of one of these new functions in use.
-
+\todo[inline]{Timo: I will create a pull request in the GeoSPARQL repository to clear up the issues concerning the spatial aggregate functions.
+Then I will update the namespaces here.
+I would find it nice to also include the functions we have discussed yesterday: transformation of SRS and in between literal formats.
+Should we mention it here as future work or should we make a draft pull request for that so that people have a reference?
+I think the latter}
 \small
@@
 
-New functions to retieve the minimum and maximum values of geometries' coordinates are also added: \emph{geof:minX} \& \emph{geof:maxX},
-\emph{geof:minY} \& \emph{geof:maxY} and \emph{geof:minZ} \& \emph{geof:maxZ}.
+New functions to retrieve the minimum and maximum values of geometries' coordinates are also added: \href{http://www.opengis.net/def/function/geosparql/minX}{\emph{geof:minX}} \& \href{http://www.opengis.net/def/function/geosparql/maxX}{\emph{geof:maxX}},
+\href{http://www.opengis.net/def/function/geosparql/minY}{\emph{geof:minY}} \& \href{http://www.opengis.net/def/function/geosparql/maxY}{\emph{geof:maxY}} and \href{http://www.opengis.net/def/function/geosparql/minZ}{\emph{geof:minZ}} \& \href{http://www.opengis.net/def/function/geosparql/maxZ}{\emph{geof:maxZ}}.
 
 \subsection{Ontology extensions}\label{sec:ontexts}
-GeoSPARQL 1.1 extends the GeoSPARQL ontology by adding a new class, \texttt{geo:SpatialMeasure}. This class represents a spatial measurement 
+GeoSPARQL 1.1 extends the GeoSPARQL ontology by adding a new class, \href{http://www.opengis.net/ont/geosparql#SpatialMeasure}{\texttt{geo:SpatialMeasure}}.
+This class represents a spatial measurement
 such as a volume, length, or area associated with a measurement amount and a unit of measure. It is the range of three newly-defined properties:
-\texttt{geo:hasArea}, \texttt{geo:hasLength} and \texttt{geo:hasVolume} which make these attributes of a geometry better accessible using 
+\href{http://www.opengis.net/ont/geosparql#hasArea}{\texttt{geo:hasArea}}, \href{http://www.opengis.net/ont/geosparql#hasLength}{\texttt{geo:hasLength}} and \href{http://www.opengis.net/ont/geosparql#hasVolume}{\texttt{geo:hasVolume}} which make these attributes of a geometry better accessible using
 SPARQL.
@@
 
-The 1.1 release also adds the property \texttt{geo:inSRS}, which allows querying for the SRS of a given geometry without the need to analyze 
+The 1.1 release also adds the property \href{http://www.opengis.net/def/function/geosparql/inSRS}{\texttt{geo:inSRS}}, which allows querying for the SRS of a given geometry without the need to analyze
 geometry serializations. The property allows for the definition of a SRS as an Internationalized Resource Identifier (IRI). It paves the way 
@@
 % Section 4
-%       new profile preentation
+%       new profile presentation
 %       new URI regimes etc

### 2021-03-12 manuscript.tex
end-to-end content, too long though

@@
 information based on the General Feature Model'' and supporting vocabularies, all for Semantic 
-Web spatial data.\\
+Web spatial data.
 
@@
 Web standard, as judged by references to it in other Semantic Web standards and its wide use in 
-Semantic Web data.\\
+Semantic Web data.
 
@@
 possible handling of new spatial referencing systems, new geometry representations, and new artifact 
-presentation.\\
+presentation.
 
@@
 \section{Motivation to update GeoSPARQL}\label{sec:motivation}
-Interest in updating GeoSPARQL had been captured\footnote{\url{https://www.w3.org/2015/spatial/wiki/Further_development_of_GeoSPARQL}}
- by the World Wide Web Consortium's (W3C) \textit{Spatial Data On The Web Working Group} (SDWWG)
-when a large body of work was done on \textit{Semantic Web} spatial data around in approximately 2015 - 2017, but 
-no updates to GeoSPARQL were ultimately made by that Working Group.
-
-% Spatial Data On The Web Best Practices \cite{van2019best}
+The Open Geospatial Consortium (OGC) \& World Wide Web Consortium's (W3C) \textit{Spatial Data On The Web Working Group}
+(SDWWG) established \textit{Spatial Data On The Web Best Practices}\cite{van_den_brink_best_2018} which noted 
+shortcomings with then current spatial data standards such as ``A best practice for returning geometries in a 
+specific requested CRS has not yet emerged''. The group also informally captured specific suggested updates for 
+GeoSPARQL\footnote{\url{https://www.w3.org/2015/spatial/wiki/Further_development_of_GeoSPARQL}}, however no updates 
+to GeoSPARQL were ultimately made by that Working Group.
 
@@
 fault fixes and proposed extensions to GeoSPARQL 1.0 are captured in an OGC White Paper \cite{geosparqlwhitepaper}. Some,
-but not all of the SDWWG's proposals are included in the White Paper with the different communities - W3C and OGC - 
-naturally reflecting different desires.
+but not all of the SDWWG's ideas have been taken up by the SDW, for example the \textit{Best Practices}~\cite{van_den_brink_best_2018}
+aspiration that ``A possible way forward is an update for the GeoSPARQL spatial ontology. This would provide an 
+agreed spatial ontology, i.e., a bridge or common ground between geographical and non-geographical spatial data...''.
+This point has been considered by the SWG and included in future releases' scope.
+
+Another \textit{Best Practices} issue raised is that ``it makes sense to publish different geometric representations 
+of a spatial object that can be used for different purposes'', again, this is being considered by the SWG with initial
+thoughts cenetering on defining \textit{roles} of geometries with respect to features.
 
@@
     \begin{itemize}
-        \item[$-$] GeoJSON and other formats that have become popular since GeoSPARQL 1.0 's publication
+        \item[$-$] GeoJSON, KML itc. Newer, popular format
     \end{itemize} 
@@
     \end{itemize} 
-    \item[$\ast$] possible alignments to other ontologies, perhaps the \textit{W3C Time Ontology in OWL}\cite{simon_cox_time_2017}
-    \item[$\ast$] catering for new and very different SRSes, such as Discrete Global Grid Systems (DGGS) 
+    \item[$\ast$] alignments to other ontologies, perhaps \textit{W3C Time Ontology in OWL}\cite{simon_cox_time_2017}
+    \item[$\ast$] catering for very different SRSes, such as Discrete Global Grid Systems
 \end{itemize}
@@
 
-An additional area of updates to GeoSPARQL that was not foreseen by GeoSPARQL 1.0, the SDWWG, or the initial proponents of the SWG
-but which was included by the early SWG attendees in GeoSPARQL 1.1 and is now present in the SWG's working repository was that 
-of new modes of standard presentation. The motivation for this was conceptual work within the W3C and the OGC to do with how 
-multi-part standards might link their parts and declare dependencies. This resulted in the profile declaration work, explained 
+An originally unforseen area of updates to GeoSPARQL is new modes of standard presentation. Motivatied by conceptual work within the W3C and the OGC for  
+multi-part standards presentation, this has resulted in profile declarations, explained 
 in the next section.
@@
 \section{Updates in GeoSPARQL 1.1}\label{sec:newfeatures}
-So far (as of June, 2020) the GeoSPARQL SWG has triaged changed requests into GeoSPARQL 1.1, 1.2 \$ 2.0 releases and has addressed
-many 1.1 requests. Here we report only on those 1.1 requests. This section describes some of the areas updated in GeoSAPRQL 1.1
-and the following section foreshadows likely 1.2 and 2.0 updates.
+So far (as of June, 2020) the GeoSPARQL SWG has triaged change requests for GeoSPARQL releases and has addressed
+many 1.1 requests which are the only ons we report here. The next sections foreshadow likely 1.2 and 2.0 updates.
 
@@
 of one of these new functions in use.
-\todo[inline]{Timo: I will create a pull request in the GeoSPARQL repository to clear up the issues concerning the spatial aggregate functions.
+% \todo[inline]{Timo: I will create a pull request in the GeoSPARQL repository to clear up the issues concerning the spatial aggregate functions.
 Then I will update the namespaces here.
@@
 \subsection{Ontology extensions}\label{sec:ontexts}
+Figure~\ref{fig:geosparql11ontology} overviews the ontology, as 
+of release 1.1.
+
 GeoSPARQL 1.1 extends the GeoSPARQL ontology by adding a new class, \href{http://www.opengis.net/ont/geosparql#SpatialMeasure}{\texttt{geo:SpatialMeasure}}.
@@
 
-The 1.1 release also adds the property \href{http://www.opengis.net/def/function/geosparql/inSRS}{\texttt{geo:inSRS}}, which allows querying for the SRS of a given geometry without the need to analyze
-geometry serializations. The property allows for the definition of a SRS as an Internationalized Resource Identifier (IRI). It paves the way 
-for a future definition of SRS fully in RDF, as anticipated for GeoSPARQL 2.0. Figure~\ref{fig:geosparql11ontology} overviews the ontology, as 
-of release 1.1.
+The 1.1 release also adds the property \href{http://www.opengis.net/def/function/geosparql/inSRS}{\texttt{geo:inSRS}}, which allows declarations of a geometry's SRS, independent of serializations.
+It paves the way
+for future definition of SRSes in RDF, as anticipated for GeoSPARQL 2.0.
 
@@
 \section{Conclusions}\label{sec:conclusions}
+An staged schedule of updates to this very important \textit{Semantic Web} spatial data standard has been initiated with simple, objvious and strictly backwards-compatable changes now in progress for GeoSPARQL 1.1.
+1.2 will include more detailed work later in 2021 and 2.0, as yet un-specified, might be a somewhat different stanandard, to be determined by more requirements gathering and feedback on the 1.1 and 1.2 releases.
 
 \subsection{Future Work}\label{sec:futurework}
-GeoSPARQL 1.2, GeoSPARQL 2.0?
 \begin{itemize}
-    \item GeoSPARQL extension ontologies: SRS systems
-    \item GeoSPARQL 1.2: More literals?
-    \item GeoSPARQL 2.0: Full featured support for 3D, simple features functions, coverages?
+    \item[$\ast$] GeoSPARQL 1.1: to be completed by July, 2021
+    \item[$\ast$] GeoSPARQL 1.2: more adventurous extension - late 2021
+    \item[$\ast$] GeoSPARQL 2.0: so-far unspecified
 \end{itemize}
-%   History of GeoSPARWL 1.0
-% Section 2
-%   motivation to update the spec
-% Section 3
-%   GeoSPARQL 1.1 additions
-%       motivation, and expected use, of new spatial functions. In particular, why in GeoSPARQL when aggregations etc. can be performed elsewhere?
-%       Spatial Measure
-% Section 4
-%       new profile presentation
-%       new URI regimes etc
-% Section 5
-%       expected changed use modes
 

### 2021-03-12 manuscript.tex
Update on Overleaf.

--- 2021-03-11 manuscript.tex
@@
 \begin{itemize}
-    \item[$\ast$] \textbf{1.1}: extensions that are fully compatable with GeoSPARQL 1.0
-    \item[$\ast$] \textbf{1.2}: fully or mostly compatable extensions but which are larger additions to the standard's conceptual coverage
-    \item[$\ast$] \textbf{2.0}: a future GeoSPARQL that might be quite different and partly incompatable with GeoSPARQL 1.0
+    \item[$\ast$] \textbf{1.1}: extensions that are fully compatible with GeoSPARQL 1.0
+    \item[$\ast$] \textbf{1.2}: fully or mostly compatible extensions but which are larger additions to the standard's conceptual coverage
+    \item[$\ast$] \textbf{2.0}: a future GeoSPARQL that might be quite different and partly incompatible with GeoSPARQL 1.0
 \end{itemize}
@@
 So far (as of June, 2020) the GeoSPARQL SWG has triaged changed requests into GeoSPARQL 1.1, 1.2 \$ 2.0 releases and has addressed
-many 1.1 requests. Here we report only on those 1.1 requests. This section describes some of the areas updated in GeoSAPRQL 1.1
+many 1.1 requests. Here we report only on those 1.1 requests. This section describes some of the areas updated in GeoSPARQL 1.1
 and the following section foreshadows likely 1.2 and 2.0 updates.
@@
 
-The profile declaration for GeosPARQL 1.0 will be published by the OGC as a stand-alone resource sometime in early 2021 along with some 
-updated GeoSPARQL 1.0 resources. The profile declaration for GeosPARQL 1.1 will be published at the same time as all, or most, of the 
+The profile declaration for GeoSPARQL 1.0 will be published by the OGC as a stand-alone resource sometime in early 2021 along with some 
+updated GeoSPARQL 1.0 resources. The profile declaration for GeoSPARQL 1.1 will be published at the same time as all, or most, of the 
 1.1 releases' updated resources, currently expected in mid-2021. The 1.1 releases' resources are:

### 2021-03-12 manuscript.tex
reduced to 7 pages

--- 2021-03-12 manuscript.tex
@@
     Nicholas J. Car\inst{1}\orcidID{0000-0002-8742-7730} \and \\
-    Timo Homburg\inst{2}\orcidID{0000-0002-9499-5840} \and \\
-    Simon J.D Cox\inst{3}\orcidID{0000-0002-3884-3420}
+    Timo Homburg\inst{2}\orcidID{0000-0002-9499-5840}\\
+%    Simon J.D Cox\inst{3}\orcidID{0000-0002-3884-3420}
 }
@@
     {
-    SURROUND Australia Pty Ltd., Australia \&\\
+    %SURROUND Australia Pty Ltd., Australia \&\\
     Australian National University, Australia\\
     \email{nicholas.car@surroundaustralia.com}\\
-    \url{https://surroundaustralia.com}
+    %\url{https://surroundaustralia.com}
     }
@@
     }
-    \and 
-    {
-        Commonwealth Scientific \& Industrial Research Organisation, Australia\\
-        \email{simon.cox@csiro.au}
-    }
+    % \and 
+    % {
+    %     Commonwealth Scientific \& Industrial Research Organisation, Australia\\
+    %     \email{simon.cox@csiro.au}
+    % }
 }
@@
 \begin{abstract}
-The Open Geospatial Consortium published the GeoSPARQL 1.0 standard in 2012 containing multiple 
-parts that define ``SPARQL extension functions'', ``RIF rules'', ``an RDF/OWL ontology for 
-information based on the General Feature Model'' and supporting vocabularies, all for Semantic 
-Web spatial data.
+In 2012 the Open Geospatial Consortium published GeoSPARQL 
+defining ``SPARQL extension functions'', ``RIF rules'', ``and RDF/OWL ontology for 
+[spatial] information'' and supporting vocabularies.\\
 
@@
 Web standard, as judged by references to it in other Semantic Web standards and its wide use in 
-Semantic Web data.
+Semantic Web data.\\
 
 An update to the standard was proposed in 2019 to deliver GeoSPARQL 1.1 in 2021 with a charter to: 
-handle outstanding change requests and source new ones from the GeoSPARQL user community as well 
+handle outstanding change requests and source new ones from the user community and 
 to ``better present'' the standard, that is to better link all the standard’s parts and better 
-document \& exemplify elements. Expected updates included possible alignments to other ontologies, 
-possible handling of new spatial referencing systems, new geometry representations, and new artifact 
-presentation.
+document \& exemplify elements. Expected updates included alignments to other ontologies, 
+handling of new spatial referencing systems, new geometry representations, and new artifact 
+presentation.\\
 
@@
 \section{Introduction}\label{sec:introduction}
-The GeoSPARQL standard, first issues in 2021 by the Open Geospatial Consortium (OGC)\footnote{\url{https://www.ogc.org}} 
+The GeoSPARQL standard, issued in 2012 by the Open Geospatial Consortium (OGC)\footnote{\url{https://www.ogc.org}} 
 is one of, if not the most\footnote{It's hard to calculate use but references to GeoSPARQL other well-known standards, 
-such as DCAT2 (\url{https://www.w3.org/TR/vocab-dcat/}) suggests this}, used \textit{Sematic Web} ontologies for 
-representing spatial data.
+such as DCAT2 (\url{https://www.w3.org/TR/vocab-dcat/}) suggests this}, used \textit{Sematic Web} standards for 
+spatial data.
 
-The original GeoSPARQL release, which we refer to as GeoSPARQL 1.0, contained a \textit{specification} document,
+The original release - GeoSPARQL 1.0 - contained a \textit{specification} document,
 a main ``GeoSPARQL'' ontology in an RDF file and a ``Simple Features Vocabulary'' ontology also in an RDF file. The 
 ``GeoSPARQL'' ontology content, as well as lists of geospatial functions that could be performed on RDF data via 
-SPARQL\footnote{\url{https://www.w3.org/TR/sparql11-query/}} queries was defined in the specification document, as
+SPARQL\footnote{\url{https://www.w3.org/TR/sparql11-query/}} queries were defined in the specification document, as
 were entailment rules and requirements \& abstract tests for testing ontology data and function implementations. 
-Within the last few years, the function lists from the specification were extracted into a SKOS\footnote{\url{https://www.w3.org/TR/skos-reference/}}
-vocabulary.
+Function lists from the specification have been extracted into SKOS\footnote{\url{https://www.w3.org/TR/skos-reference/}}
+vocabularies.
 
 \section{Motivation to update GeoSPARQL}\label{sec:motivation}
-The Open Geospatial Consortium (OGC) \& World Wide Web Consortium's (W3C) \textit{Spatial Data On The Web Working Group}
+The OGC \& World Wide Web Consortium's (W3C) \textit{Spatial Data On The Web Working Group}
 (SDWWG) established \textit{Spatial Data On The Web Best Practices}\cite{van_den_brink_best_2018} which noted 
-shortcomings with then current spatial data standards such as ``A best practice for returning geometries in a 
+shortcomings with then current spatial data standards: ``A best practice for returning geometries in a 
 specific requested CRS has not yet emerged''. The group also informally captured specific suggested updates for 
 GeoSPARQL\footnote{\url{https://www.w3.org/2015/spatial/wiki/Further_development_of_GeoSPARQL}}, however no updates 
-to GeoSPARQL were ultimately made by that Working Group.
+to GeoSPARQL were then made.
 
@@
 fault fixes and proposed extensions to GeoSPARQL 1.0 are captured in an OGC White Paper \cite{geosparqlwhitepaper}. Some,
-but not all of the SDWWG's ideas have been taken up by the SDW, for example the \textit{Best Practices}~\cite{van_den_brink_best_2018}
+but not all, of the SDWWG's ideas have been taken up by the SDW, for example the \textit{Best Practices}~\cite{van_den_brink_best_2018}
 aspiration that ``A possible way forward is an update for the GeoSPARQL spatial ontology. This would provide an 
@@
 Another \textit{Best Practices} issue raised is that ``it makes sense to publish different geometric representations 
-of a spatial object that can be used for different purposes'', again, this is being considered by the SWG with initial
-thoughts cenetering on defining \textit{roles} of geometries with respect to features.
+of a spatial object that can be used for different purposes''. This is being considered by the SWG with initial
+thoughts centering on defining \textit{roles} of geometries with respect to features.
 
@@
 
-The first two of these new serializations have been much anticipated and were requested by the SDWWG and many users of 
+GeoJSON \& KML have been much anticipated and were requested by the SDWWG and many users of 
 GeoSPARQL, due to those formats' popularity.
 
-The third new serialization format is more forward-looking in that it is not driven by user demand but by predicted demand.
+The DGGS format is more forward-looking in that it is not driven by user demand but by predicted demand.
 DGGS does not have a single, concrete format standard as the others do, nor is it ever likely to - different DGGSes will 
@@
 
-New functions to retrieve the minimum and maximum values of geometries' coordinates are also added: \href{http://www.opengis.net/def/function/geosparql/minX}{\emph{geof:minX}} \& \href{http://www.opengis.net/def/function/geosparql/maxX}{\emph{geof:maxX}},
+Functions to retrieve min/max values of geometries' coordinates are added: \href{http://www.opengis.net/def/function/geosparql/minX}{\emph{geof:minX}} \& \href{http://www.opengis.net/def/function/geosparql/maxX}{\emph{geof:maxX}},
 \href{http://www.opengis.net/def/function/geosparql/minY}{\emph{geof:minY}} \& \href{http://www.opengis.net/def/function/geosparql/maxY}{\emph{geof:maxY}} and \href{http://www.opengis.net/def/function/geosparql/minZ}{\emph{geof:minZ}} \& \href{http://www.opengis.net/def/function/geosparql/maxZ}{\emph{geof:maxZ}}.
@@
 \subsection{Ontology extensions}\label{sec:ontexts}
-Figure~\ref{fig:geosparql11ontology} overviews the ontology, as 
-of release 1.1.
-
-GeoSPARQL 1.1 extends the GeoSPARQL ontology by adding a new class, \href{http://www.opengis.net/ont/geosparql#SpatialMeasure}{\texttt{geo:SpatialMeasure}}.
+GeoSPARQL 1.1 - see Figure~\ref{fig:geosparql11ontology} for an overview - extends the GeoSPARQL ontology by adding a new class, \href{http://www.opengis.net/ont/geosparql#SpatialMeasure}{\texttt{geo:SpatialMeasure}}.
 This class represents a spatial measurement
@@
 
-These additions address a number of change requests from the SDWWG and the SWG but, given the introduction of a dedicated \textit{measure} class, 
-the do more than just provide for a few scalar spatial measures and open up GeoSPARQL to general patterns of measurement present in ontologies 
+These additions address a several requests from the SDWWG \& SWG but, given the introduction of a dedicated \textit{measure} class, 
+do more too: they open up GeoSPARQL to general patterns of measurement present in ontologies 
 such as the W3C's \textit{SOSA}~\cite{haller_semantic_2017}.
 
-The 1.1 release also adds the property \href{http://www.opengis.net/def/function/geosparql/inSRS}{\texttt{geo:inSRS}}, which allows declarations of a geometry's SRS, independent of serializations.
-It paves the way
+The 1.1 release also adds the property \href{http://www.opengis.net/def/function/geosparql/inSRS}{\texttt{geo:inSRS}}, which allows declarations of a geometry's SRS, independent of serializations which paves the way
 for future definition of SRSes in RDF, as anticipated for GeoSPARQL 2.0.
@@
 \section{Conclusions}\label{sec:conclusions}
-An staged schedule of updates to this very important \textit{Semantic Web} spatial data standard has been initiated with simple, objvious and strictly backwards-compatable changes now in progress for GeoSPARQL 1.1.
-1.2 will include more detailed work later in 2021 and 2.0, as yet un-specified, might be a somewhat different stanandard, to be determined by more requirements gathering and feedback on the 1.1 and 1.2 releases.
+A staged schedule of updates to this important \textit{Semantic Web} spatial standard has been initiated with simple and strictly backwards-compatable changes now in GeoSPARQL 1.1.
+1.2 will include more detailed work later in 2021 and 2.0, as yet un-specified, might be somewhat different: to be determined by more requirements gathering and feedback on the 1.1 and 1.2 releases.
 

### 2021-03-13 manuscript.tex
further shortening

@@
     Nicholas J. Car\inst{1}\orcidID{0000-0002-8742-7730} \and \\
-    Timo Homburg\inst{2}\orcidID{0000-0002-9499-5840}\\
+    Timo Homburg\inst{2}\orcidID{0000-0002-9499-5840}
 %    Simon J.D Cox\inst{3}\orcidID{0000-0002-3884-3420}
@@
 The OGC \& World Wide Web Consortium's (W3C) \textit{Spatial Data On The Web Working Group}
-(SDWWG) established \textit{Spatial Data On The Web Best Practices}\cite{van_den_brink_best_2018} which noted 
+(SDWWG) established \textit{Spatial Data On The Web Best Practices}~\cite{van_den_brink_best_2018} which noted 
 shortcomings with then current spatial data standards: ``A best practice for returning geometries in a 
@@
 motivation for work within the area of GeoSPARQL, that of \textit{Semantic Web} spatial data, and a series of
-fault fixes and proposed extensions to GeoSPARQL 1.0 are captured in an OGC White Paper \cite{geosparqlwhitepaper}. Some,
+fault fixes and proposed extensions to GeoSPARQL 1.0 are captured in an OGC White Paper~\cite{geosparqlwhitepaper}. Some,
 but not all, of the SDWWG's ideas have been taken up by the SDW, for example the \textit{Best Practices}~\cite{van_den_brink_best_2018}
@@
 
-The SWG's charter - its final scope of work - is also published by the OGC \cite{abhayaratna2020ogc} and this guides 
+The SWG's charter - its final scope of work - is also published by the OGC~\cite{abhayaratna2020ogc} and this guides 
 the SWG's activities. Specific actions of the SWG and their staging are explained through the use of a publicly-available 
@@
     \end{itemize} 
-    \item[$\ast$] alignments to other ontologies, perhaps \textit{W3C Time Ontology in OWL}\cite{simon_cox_time_2017}
+    \item[$\ast$] alignments to other ontologies, perhaps \textit{W3C Time Ontology in OWL}~\cite{simon_cox_time_2017}
     \item[$\ast$] catering for very different SRSes, such as Discrete Global Grid Systems
@@
 One of the first actions undertaken by the SWG was to link the GeoSPARQL 1.0 elements through a \textit{profile} 
-declaration, where a profile is a special type of \textit{specification}, as defined by \textit{The Profiles Vocabulary}\cite{atkinson_profiles_2020}. 
+declaration, where a profile is a special type of \textit{specification}, as defined by \textit{The Profiles Vocabulary}~\cite{atkinson_profiles_2020}. 
 The specific motivation for this was the SWG's recognition that GeoSPARQL 1.0 consisted of multiple parts, not all
@@
 The profile declaration for GeoSPARQL 1.0 will be published by the OGC as a stand-alone resource sometime in early 2021 along with some 
-updated GeoSPARQL 1.0 resources. The profile declaration for GeoSPARQL 1.1 will be published at the same time as all, or most, of the 
-1.1 releases' updated resources, currently expected in mid-2021. The 1.1 releases' resources are:
+updated GeoSPARQL 1.0 resources. The profile declaration for GeoSPARQL 1.1 will be published at the same time as the 
+1.1 releases, currently expected in mid-2021. All 1.0 and 1.1 relase resources are currently available in draft form 
+in the SWG's online code repository\footnote{\url{https://github.com/opengeospatial/ogc-geosparql}}. The 1.1 releases' resources are:
 
@@
     \item a Simple Features feature types vocabulary
-    \item a set of RIF\cite{kifer2013rif} rules
-    \item SHACL\cite{knublauch_shapes_2017} shapes for RDF data validation 
+    \item a set of RIF~\cite{kifer2013rif} rules
+    \item SHACL~\cite{knublauch_shapes_2017} shapes for RDF data validation 
 \end{enumerate}
-
-Both the 1.0 and the 1.1 resources are currently available from the SWG's online code repository\footnote{\url{https://github.com/opengeospatial/ogc-geosparql}}.
 
@@
 \begin{enumerate}
-    \item \textbf{GeoJSON} (Geo- JavaScript Object Notation)\cite{butler2016geojson}
-    \item \textbf{KML} (Keyhole Markup Language)\cite{nolan2014keyhole} 
-    \item \textbf{DGGS} (Discrete Global Grid System)\cite{sahr1998discrete}
+    \item \textbf{GeoJSON} (Geo- JavaScript Object Notation)~\cite{butler2016geojson}
+    \item \textbf{KML} (Keyhole Markup Language)~\cite{nolan2014keyhole} 
+    \item \textbf{DGGS} (Discrete Global Grid System)~\cite{sahr1998discrete}
 \end{enumerate}
@@
     \item No client-side library is needed to create an aggregated geometry result
-    \item Fewer and more appropriate results can be returned using a spatial aggregate function (for example, a union geometry result)
-    \item Spatial aggregates from different SPARQL endpoints can easily be calculated using federated queries 
+    \item Fewer/more appropriate results returned, for example a union result
+    \item Federated SPARQL queries can aggregate results from multiple endpoints
 \end{enumerate}
 
-In addition to the \href{http://www.opengis.net/def/function/geosparql/union}{\emph{geof:union}}, \href{http://www.opengis.net/def/function/geosparql/envelope}{\emph{geof:envelope}} and \href{http://www.opengis.net/def/function/geosparql/convexHull}{\emph{geof:convexHull}} functions already defined in GeoSPARQL 1.0,
-for use within SPARQL \texttt{FILTER} operations, GeoSPARQL 1.1 defines \emph{geosf:Union}, \emph{geosf:Envelope} and \emph{geosf:ConvexHull}
+In addition to \href{http://www.opengis.net/def/function/geosparql/union}{\emph{geof:union}}, \href{http://www.opengis.net/def/function/geosparql/envelope}{\emph{geof:envelope}} and \href{http://www.opengis.net/def/function/geosparql/convexHull}{\emph{geof:convexHull}} defined in GeoSPARQL 1.0
+for use within SPARQL \texttt{FILTER} operations, 1.1 defines \emph{geosf:Union}, \emph{geosf:Envelope} and \emph{geosf:ConvexHull}
 as well as \emph{geosf:BoundingCircle}, \emph{geosf:Centroid}, \emph{geosf:ConcatLines} - concatenating a set of overlapping linestrings 
-that overlap - and \emph{geosf:ConcaveHull} that can be used to return aggregated results. Listing~\ref{lst:agfunc} gives an example 
+that overlap - and \emph{geosf:ConcaveHull} that can return aggregated results. Listing~\ref{lst:agfunc} gives an example 
 of one of these new functions in use.
@@
 
-These additions address a several requests from the SDWWG \& SWG but, given the introduction of a dedicated \textit{measure} class, 
+These additions address requests from the SDWWG \& SWG but
 do more too: they open up GeoSPARQL to general patterns of measurement present in ontologies 
-such as the W3C's \textit{SOSA}~\cite{haller_semantic_2017}.
-
-The 1.1 release also adds the property \href{http://www.opengis.net/def/function/geosparql/inSRS}{\texttt{geo:inSRS}}, which allows declarations of a geometry's SRS, independent of serializations which paves the way
-for future definition of SRSes in RDF, as anticipated for GeoSPARQL 2.0.
+such as the W3C's \textit{SOSA}~\cite{haller_semantic_2017}. Similarly, 
+the 1.1 release addition of property \href{http://www.opengis.net/def/function/geosparql/inSRS}{\texttt{geo:inSRS}}, allows declarations of a geometry's SRS, independent of serializations and paves the way
+for future definition of SRSes in RDF, anticipated for GeoSPARQL 2.0.
 

### 2021-03-15 manuscript.tex
Update on Overleaf.

@@
 In 2012 the Open Geospatial Consortium published GeoSPARQL 
-defining ``SPARQL extension functions'', ``RIF rules'', ``and RDF/OWL ontology for 
-[spatial] information'' and supporting vocabularies.\\
+defining ``SPARQL extension functions'', ``RIF rules'', ``an RDF/OWL ontology for [spatial] information'' and supporting vocabularies.\\
 
@@
 is one of, if not the most\footnote{It's hard to calculate use but references to GeoSPARQL other well-known standards, 
-such as DCAT2 (\url{https://www.w3.org/TR/vocab-dcat/}) suggests this}, used \textit{Sematic Web} standards for 
+such as DCAT2 (\url{https://www.w3.org/TR/vocab-dcat/}) or CIDOC-CRM \cite{doerr2013crmgeo} suggests this}, used \textit{Semantic Web} standards for 
 spatial data.
@@
 
-Recently, 2019, the OGC reconstituted a \textit{GeoSPARQL Standards Working Group} (SWG) to update GeoSPARQL. The general 
+Recently, in 2019, the OGC reconstituted a \textit{GeoSPARQL Standards Working Group} (SWG) to update GeoSPARQL. The general 
 motivation for work within the area of GeoSPARQL, that of \textit{Semantic Web} spatial data, and a series of
@@
     \begin{itemize}
-        \item[$-$] GeoJSON, KML itc. Newer, popular format
+        \item[$-$] GeoJSON, KML etc. Newer, popular formats
     \end{itemize} 
@@
 So far (as of June, 2020) the GeoSPARQL SWG has triaged change requests for GeoSPARQL releases and has addressed
-many 1.1 requests which are the only ons we report here. The next sections foreshadow likely 1.2 and 2.0 updates.
+many 1.1 requests which are the only ones we report here. The next sections foreshadow likely 1.2 and 2.0 updates.
 
@@
 updated GeoSPARQL 1.0 resources. The profile declaration for GeoSPARQL 1.1 will be published at the same time as the 
-1.1 releases, currently expected in mid-2021. All 1.0 and 1.1 relase resources are currently available in draft form 
+1.1 releases, currently expected in mid-2021. All 1.0 and 1.1 release resources are currently available in draft form 
 in the SWG's online code repository\footnote{\url{https://github.com/opengeospatial/ogc-geosparql}}. The 1.1 releases' resources are:
@@
 \subsection{New spatial functions}\label{sec:newfunctions}
-While spatial aggregation functions are the norm in many non-semantic geospatial databases such as POSTGIS or Oracle Spatial, 
+While spatial aggregation functions are the norm in many non-semantic geospatial databases such as PostGIS or Oracle Spatial, 
 at the time of defining the GeoSPARQL 1.0 standard, aggregation functions had not yet been introduced into the SPARQL standard,
-but have been since with SPARQL 1.1~\cite{w3c_sparql_working_group_sparql_2013}. Spatial aggregation functions 
+but have been with SPARQL 1.1~\cite{w3c_sparql_working_group_sparql_2013}. Spatial aggregation functions 
 similar to traditional (relation database), aggregation functions such as AVG, MAX, or MIN allow aggregated results of geometry 
@@
     \includegraphics[width=\linewidth]{images/geold_ontology.png}
-    \caption{GeoSPARQL 1.1 ontology including one example feature}
+    \caption{Excerpt of the GeoSPARQL 1.1 ontology including one example feature}
     \label{fig:geosparql11ontology}
@@
 \section{Conclusions}\label{sec:conclusions}
-A staged schedule of updates to this important \textit{Semantic Web} spatial standard has been initiated with simple and strictly backwards-compatable changes now in GeoSPARQL 1.1.
-1.2 will include more detailed work later in 2021 and 2.0, as yet un-specified, might be somewhat different: to be determined by more requirements gathering and feedback on the 1.1 and 1.2 releases.
-
-\subsection{Future Work}\label{sec:futurework}
-\begin{itemize}
-    \item[$\ast$] GeoSPARQL 1.1: to be completed by July, 2021
-    \item[$\ast$] GeoSPARQL 1.2: more adventurous extension - late 2021
-    \item[$\ast$] GeoSPARQL 2.0: so-far unspecified
-\end{itemize}
+A staged schedule of updates to this important \textit{Semantic Web} spatial standard has been initiated with simple and strictly backwards-compatible changes now in GeoSPARQL 1.1.
+Features discussed for GeoSPARQL 1.2 include the formalization of coordinate reference systems in RDF, the depiction of accurracies and level of detail and the addition of further - possibly also binary - literal types.
+Work on GeoSPARQL 1.2 will start later in 2021.
+GeoSPARQL 2.0, as yet un-specified, is likely to introduce more substantial changes to the standard.
+Changes proposed for GeoSPARQL 2.0 include to broaden the scope of GeoSPARQL to further kinds of spatial data.
+To that end, full-featured support for 3D geometries and support for coverages are discussed on the level of data representations.
+These proposals are related to some growing interest in the semantic web community in representing further geospatial data related to building modeling information \cite{zhang2018bimsparql} and coverage data \cite{homburg2020geosparql+}.
+More requirements might also be introduced once feedback has been received from the GeoSPARQL 1.1 and 1.2 releases.
 

### 2021-03-18 manuscript.tex
some of Matthais' comments addressed

@@
 The GeoSPARQL standard, issued in 2012 by the Open Geospatial Consortium (OGC)\footnote{\url{https://www.ogc.org}} 
-is one of, if not the most\footnote{It's hard to calculate use but references to GeoSPARQL other well-known standards, 
-such as DCAT2 (\url{https://www.w3.org/TR/vocab-dcat/}) or CIDOC-CRM \cite{doerr2013crmgeo} suggests this}, used \textit{Semantic Web} standards for 
-spatial data.
+is one of, if not the most, popular \textit{Semantic Web} standards for spatial data.\footnote{References to GeoSPARQL other well-known standards, 
+such as DCAT2 (\url{https://www.w3.org/TR/vocab-dcat/}) or CIDOC-CRM \cite{doerr2013crmgeo} suggests this, as do the incoming links in Linked Open Vocabularies \url{https://lov.linkeddata.es/dataset/lov/vocabs/gsp} and the list of implementers that includes most of the popular triplestore vendors, a list of which has been compiled here: \url{https://github.com/opengeospatial/ogc-geosparql/issues/59}}
 

### 2021-04-17 manuscript.tex
Corrected minor remarks of the reviewers

@@
 
+In this publication we discuss the motivations behind updating GeoSPARQL 1.0 in \Cref{sec:motivation}, the content of the planned updates for the upcoming GeoSPARQL 1.1 release in \Cref{sec:newfeatures} and finally in \Cref{sec:conclusions} provide an outlook to further feature requests which are likely to be tackled in future GeoSPARQL 1.2 and 2.0 releases.
+
 \section{Motivation to update GeoSPARQL}\label{sec:motivation}
@@
     \begin{itemize}
-        \item[$-$] GeoJSON, KML etc. Newer, popular formats
+        \item[$-$] GeoJSON, KML and other popular formats
     \end{itemize} 
@@
 but look to generalised mechanisms for describing dimensions of features of which \textit{geometry} is one of many (and perhaps)
-\textit{temporality} too).
+\textit{temporality} too.
 
@@
 In addition to \href{http://www.opengis.net/def/function/geosparql/union}{\emph{geof:union}}, \href{http://www.opengis.net/def/function/geosparql/envelope}{\emph{geof:envelope}} and \href{http://www.opengis.net/def/function/geosparql/convexHull}{\emph{geof:convexHull}} defined in GeoSPARQL 1.0
-for use within SPARQL \texttt{FILTER} operations, 1.1 defines \emph{geosf:Union}, \emph{geosf:Envelope} and \emph{geosf:ConvexHull}
-as well as \emph{geosf:BoundingCircle}, \emph{geosf:Centroid}, \emph{geosf:ConcatLines} - concatenating a set of overlapping linestrings 
+for use within SPARQL \texttt{FILTER} operations, 1.1 defines \href{http://www.opengis.net/def/function/geosparql/union2}{\emph{geof:union2}} and 
+as well as \href{http://www.opengis.net/def/function/geosparql/boundingCircle}{\emph{geof:boundingCircle}}, \href{http://www.opengis.net/def/function/geosparql/centroid}{\emph{geof:centroid}}, \href{http://www.opengis.net/def/function/geosparql/concatLines}{\emph{geof:ConcatLines}} - concatenating a set of overlapping linestrings
 that overlap - and \emph{geosf:ConcaveHull} that can return aggregated results. Listing~\ref{lst:agfunc} gives an example 
@@
 # of Feature <x>
-SELECT (geosaf:BoundingCircle(?geo) AS ?circ)
+SELECT (geof:BoundingCircle(?geo) AS ?circ)
 WHERE {
@@
 
-\section{Conclusions}\label{sec:conclusions}
+\section{Conclusions and Outlook}\label{sec:conclusions}
 A staged schedule of updates to this important \textit{Semantic Web} spatial standard has been initiated with simple and strictly backwards-compatible changes now in GeoSPARQL 1.1.

### 2021-04-17 manuscript.tex
Added paragraph discussing OGC API Features

@@
 but have been with SPARQL 1.1~\cite{w3c_sparql_working_group_sparql_2013}. Spatial aggregation functions 
-similar to traditional (relation database), aggregation functions such as AVG, MAX, or MIN allow aggregated results of geometry 
+similar to traditional (relational database), aggregation functions such as AVG, MAX, or MIN allow aggregated results of geometry 
 queries, for example, to create the union of a set of selected serialized geometries. While calculating these aggregates is 
@@
 of one of these new functions in use.
+
 % \todo[inline]{Timo: I will create a pull request in the GeoSPARQL repository to clear up the issues concerning the spatial aggregate functions.
@@
 \href{http://www.opengis.net/def/function/geosparql/minY}{\emph{geof:minY}} \& \href{http://www.opengis.net/def/function/geosparql/maxY}{\emph{geof:maxY}} and \href{http://www.opengis.net/def/function/geosparql/minZ}{\emph{geof:minZ}} \& \href{http://www.opengis.net/def/function/geosparql/maxZ}{\emph{geof:maxZ}}.
+
+In parallel, the OGC API Features standard \cite{ogcapifeaturespart3} is developed, which also proposes functions for filtering feature collections.
+Even though, OGC API Features proposes the use of the Common Query Language (CQL) for filtering, it is open to other query language implementations such as GeoSPARQL.
+When comparing the filter capabilities of CQL to GeoSPARQL, one can observe that concerning spatial filter capabilities, the two query languages provide comparable functionality.
+However, the version of CQL proposed by OGC API Features also provides support of spatiotemporal operators, which may be an addition to be further explored in the contiuous GeoSPARQL development process.
 

### 2021-04-18 manuscript.tex
Added reference to the living document

@@
 So far (as of June, 2020) the GeoSPARQL SWG has triaged change requests for GeoSPARQL releases and has addressed
-many 1.1 requests which are the only ones we report here. The next sections foreshadow likely 1.2 and 2.0 updates.
+many 1.1 requests which are the only ones we report here and which are accessible through the living document for the GeoSPARQL 1.1 standard \cite{ogcgeosparql11}.
+The next sections foreshadow likely 1.2 and 2.0 updates.
 

### 2021-04-21 manuscript.tex
tiny wording change

@@
 In parallel, the OGC API Features standard \cite{ogcapifeaturespart3} is developed, which also proposes functions for filtering feature collections.
-Even though, OGC API Features proposes the use of the Common Query Language (CQL) for filtering, it is open to other query language implementations such as GeoSPARQL.
+While OGC API Features proposes the use of the Common Query Language (CQL) for filtering, it is also open to other query language implementations such as GeoSPARQL.
 When comparing the filter capabilities of CQL to GeoSPARQL, one can observe that concerning spatial filter capabilities, the two query languages provide comparable functionality.

### 2021-04-21 manuscript.tex
Nick's edits for reviewers

--- 2021-04-17 manuscript.tex
@@
 The GeoSPARQL standard, issued in 2012 by the Open Geospatial Consortium (OGC)\footnote{\url{https://www.ogc.org}} 
-is one of, if not the most, popular \textit{Semantic Web} standards for spatial data.\footnote{References to GeoSPARQL other well-known standards, 
-such as DCAT2 (\url{https://www.w3.org/TR/vocab-dcat/}) or CIDOC-CRM \cite{doerr2013crmgeo} suggests this, as do the incoming links in Linked Open Vocabularies \url{https://lov.linkeddata.es/dataset/lov/vocabs/gsp} and the list of implementers that includes most of the popular triplestore vendors, a list of which has been compiled here: \url{https://github.com/opengeospatial/ogc-geosparql/issues/59}}
+is one of, if not the most, popular \textit{Semantic Web} standards for spatial data.\footnote{References to GeoSPARQL in other well-known standards, 
+such as DCAT2 (\url{https://www.w3.org/TR/vocab-dcat/}) and CIDOC-CRM \cite{doerr2013crmgeo} suggests it is popular, as do the incoming links in Linked Open Vocabularies \url{https://lov.linkeddata.es/dataset/lov/vocabs/gsp} and the list of implementers that includes most of the popular triplestore vendors, a list of which has been compiled here: \url{https://github.com/opengeospatial/ogc-geosparql/issues/59}}
 
@@
 vocabularies.
-
-In this publication we discuss the motivations behind updating GeoSPARQL 1.0 in \Cref{sec:motivation}, the content of the planned updates for the upcoming GeoSPARQL 1.1 release in \Cref{sec:newfeatures} and finally in \Cref{sec:conclusions} provide an outlook to further feature requests which are likely to be tackled in future GeoSPARQL 1.2 and 2.0 releases.
 
@@
 \begin{itemize}
-    \item[$\ast$] new geometry serializations
+    \item new geometry serializations
     \begin{itemize}
-        \item[$-$] GeoJSON, KML and other popular formats
+        \item[$-$] GeoJSON, KML etc. Newer, popular formats
     \end{itemize} 
-    \item[$\ast$] new ontology classes to cater for more nuanced spatial information
-    \item[$\ast$] more spatial functions
+    \item new ontology classes to cater for more nuanced spatial information
+    \item more spatial functions
     \begin{itemize}
@@
     \end{itemize} 
-    \item[$\ast$] scalar spatial properties 
+    \item scalar spatial properties 
     \begin{itemize}
@@
     \end{itemize} 
-    \item[$\ast$] better handling of Spatial (Coordinate) Reference Systems (SRS)
+    \item better handling of Spatial (Coordinate) Reference Systems (SRS)
     \begin{itemize}
@@
     \end{itemize} 
-    \item[$\ast$] Internet protocol-based selection of different geometries for features
+    \item Internet protocol-based selection of different geometries for features
 \end{itemize}
@@
 \begin{itemize}
-    \item[$\ast$] a revision of the ``upper ontology'' structuring of GeoSPARQL
+    \item a revision of the ``upper ontology'' structuring of GeoSPARQL
     \begin{itemize}
@@
     \end{itemize} 
-    \item[$\ast$] alignments to other ontologies, perhaps \textit{W3C Time Ontology in OWL}~\cite{simon_cox_time_2017}
-    \item[$\ast$] catering for very different SRSes, such as Discrete Global Grid Systems
+    \item alignments to other ontologies, perhaps \textit{W3C Time Ontology in OWL}~\cite{simon_cox_time_2017}
+    \item catering for very different SRSes, such as Discrete Global Grid Systems
 \end{itemize}
@@
 \begin{itemize}
-    \item[$\ast$] \textbf{1.1}: extensions that are fully compatible with GeoSPARQL 1.0
-    \item[$\ast$] \textbf{1.2}: fully or mostly compatible extensions but which are larger additions to the standard's conceptual coverage
-    \item[$\ast$] \textbf{2.0}: a future GeoSPARQL that might be quite different and partly incompatible with GeoSPARQL 1.0
+    \item \textbf{1.1}: extensions that are fully compatible with GeoSPARQL 1.0
+    \item \textbf{1.2}: fully or mostly compatible extensions but which are larger additions to the standard's conceptual coverage
+    \item \textbf{2.0}: a future GeoSPARQL that might be quite different and partly incompatible with GeoSPARQL 1.0
 \end{itemize}
@@
 move away from only, or primarily, \textit{geo}spatial, or perhaps focus not just on \texttt{Feature}/\texttt{Geometry} relations
-but look to generalised mechanisms for describing dimensions of features of which \textit{geometry} is one of many (and perhaps)
-\textit{temporality} too.
+but look to generalised mechanisms for describing dimensions of features of which \textit{geometry} is just one of many, and 
+\textit{temporality} might be another.
 
@@
 \section{Updates in GeoSPARQL 1.1}\label{sec:newfeatures}
-So far (as of June, 2020) the GeoSPARQL SWG has triaged change requests for GeoSPARQL releases and has addressed
+So far (as of April, 2021) the GeoSPARQL SWG has triaged change requests for GeoSPARQL releases and has addressed
 many 1.1 requests which are the only ones we report here. The next sections foreshadow likely 1.2 and 2.0 updates.
@@
 but have been with SPARQL 1.1~\cite{w3c_sparql_working_group_sparql_2013}. Spatial aggregation functions 
-similar to traditional (relational database), aggregation functions such as AVG, MAX, or MIN allow aggregated results of geometry 
+similar to traditional (relation database), aggregation functions such as AVG, MAX, or MIN allow aggregated results of geometry 
 queries, for example, to create the union of a set of selected serialized geometries. While calculating these aggregates is 
@@
 In addition to \href{http://www.opengis.net/def/function/geosparql/union}{\emph{geof:union}}, \href{http://www.opengis.net/def/function/geosparql/envelope}{\emph{geof:envelope}} and \href{http://www.opengis.net/def/function/geosparql/convexHull}{\emph{geof:convexHull}} defined in GeoSPARQL 1.0
-for use within SPARQL \texttt{FILTER} operations, 1.1 defines \href{http://www.opengis.net/def/function/geosparql/union2}{\emph{geof:union2}} and 
-as well as \href{http://www.opengis.net/def/function/geosparql/boundingCircle}{\emph{geof:boundingCircle}}, \href{http://www.opengis.net/def/function/geosparql/centroid}{\emph{geof:centroid}}, \href{http://www.opengis.net/def/function/geosparql/concatLines}{\emph{geof:ConcatLines}} - concatenating a set of overlapping linestrings
+for use within SPARQL \texttt{FILTER} operations, 1.1 defines \emph{geosf:Union}, \emph{geosf:Envelope} and \emph{geosf:ConvexHull}
+as well as \emph{geosf:BoundingCircle}, \emph{geosf:Centroid}, \emph{geosf:ConcatLines} - concatenating a set of overlapping linestrings 
 that overlap - and \emph{geosf:ConcaveHull} that can return aggregated results. Listing~\ref{lst:agfunc} gives an example 
 of one of these new functions in use.
-
 % \todo[inline]{Timo: I will create a pull request in the GeoSPARQL repository to clear up the issues concerning the spatial aggregate functions.
@@
 # of Feature <x>
-SELECT (geof:BoundingCircle(?geo) AS ?circ)
+SELECT (geosaf:BoundingCircle(?geo) AS ?circ)
 WHERE {
@@
 \href{http://www.opengis.net/def/function/geosparql/minY}{\emph{geof:minY}} \& \href{http://www.opengis.net/def/function/geosparql/maxY}{\emph{geof:maxY}} and \href{http://www.opengis.net/def/function/geosparql/minZ}{\emph{geof:minZ}} \& \href{http://www.opengis.net/def/function/geosparql/maxZ}{\emph{geof:maxZ}}.
-
-In parallel, the OGC API Features standard \cite{ogcapifeaturespart3} is developed, which also proposes functions for filtering feature collections.
-Even though, OGC API Features proposes the use of the Common Query Language (CQL) for filtering, it is open to other query language implementations such as GeoSPARQL.
-When comparing the filter capabilities of CQL to GeoSPARQL, one can observe that concerning spatial filter capabilities, the two query languages provide comparable functionality.
-However, the version of CQL proposed by OGC API Features also provides support of spatiotemporal operators, which may be an addition to be further explored in the contiuous GeoSPARQL development process.
 
@@
 
-\section{Conclusions and Outlook}\label{sec:conclusions}
+\section{Conclusions}\label{sec:conclusions}
 A staged schedule of updates to this important \textit{Semantic Web} spatial standard has been initiated with simple and strictly backwards-compatible changes now in GeoSPARQL 1.1.

### 2021-04-21 manuscript.tex

@@
 So far (as of April, 2021) the GeoSPARQL SWG has triaged change requests for GeoSPARQL releases and has addressed
-many 1.1 requests which are the only ones we report here. The next sections foreshadow likely 1.2 and 2.0 updates.
+many 1.1 requests which are the only ones we report here and which are accessible through the living document for the GeoSPARQL 1.1 standard \cite{ogcgeosparql11}.
+The next sections foreshadow likely 1.2 and 2.0 updates.
 
@@
 In addition to \href{http://www.opengis.net/def/function/geosparql/union}{\emph{geof:union}}, \href{http://www.opengis.net/def/function/geosparql/envelope}{\emph{geof:envelope}} and \href{http://www.opengis.net/def/function/geosparql/convexHull}{\emph{geof:convexHull}} defined in GeoSPARQL 1.0
-for use within SPARQL \texttt{FILTER} operations, 1.1 defines \emph{geosf:Union}, \emph{geosf:Envelope} and \emph{geosf:ConvexHull}
-as well as \emph{geosf:BoundingCircle}, \emph{geosf:Centroid}, \emph{geosf:ConcatLines} - concatenating a set of overlapping linestrings 
+for use within SPARQL \texttt{FILTER} operations, 1.1 defines \href{http://www.opengis.net/def/function/geosparql/union2}{\emph{geof:union2}} and 
+as well as \href{http://www.opengis.net/def/function/geosparql/boundingCircle}{\emph{geof:boundingCircle}}, \href{http://www.opengis.net/def/function/geosparql/centroid}{\emph{geof:centroid}}, \href{http://www.opengis.net/def/function/geosparql/concatLines}{\emph{geof:ConcatLines}} - concatenating a set of overlapping linestrings
 that overlap - and \emph{geosf:ConcaveHull} that can return aggregated results. Listing~\ref{lst:agfunc} gives an example 
 of one of these new functions in use.
+
 % \todo[inline]{Timo: I will create a pull request in the GeoSPARQL repository to clear up the issues concerning the spatial aggregate functions.
@@
 \href{http://www.opengis.net/def/function/geosparql/minY}{\emph{geof:minY}} \& \href{http://www.opengis.net/def/function/geosparql/maxY}{\emph{geof:maxY}} and \href{http://www.opengis.net/def/function/geosparql/minZ}{\emph{geof:minZ}} \& \href{http://www.opengis.net/def/function/geosparql/maxZ}{\emph{geof:maxZ}}.
+
+In parallel, the OGC API Features standard \cite{ogcapifeaturespart3} is developed, which also proposes functions for filtering feature collections.
+While OGC API Features proposes the use of the Common Query Language (CQL) for filtering, it is also open to other query language implementations such as GeoSPARQL.
+When comparing the filter capabilities of CQL to GeoSPARQL, one can observe that concerning spatial filter capabilities, the two query languages provide comparable functionality.
+However, the version of CQL proposed by OGC API Features also provides support of spatiotemporal operators, which may be an addition to be further explored in the contiuous GeoSPARQL development process.
 

### 2021-04-21 manuscript.tex
BibTek improvements

@@
     {
-    %SURROUND Australia Pty Ltd., Australia \&\\
+    SURROUND Australia Pty Ltd., Australia \&\\
     Australian National University, Australia\\
     \email{nicholas.car@surroundaustralia.com}\\
-    %\url{https://surroundaustralia.com}
     }
@@
     }
-    % \and 
-    % {
-    %     Commonwealth Scientific \& Industrial Research Organisation, Australia\\
-    %     \email{simon.cox@csiro.au}
-    % }
 }
@@
 In 2012 the Open Geospatial Consortium published GeoSPARQL 
-defining ``SPARQL extension functions'', ``RIF rules'', ``an RDF/OWL ontology for [spatial] information'' and supporting vocabularies.\\
+defining ``SPARQL extension functions'', ``RIF rules'' and ``an RDF/OWL ontology for [spatial] information''.\\
 

### 2021-04-21 manuscript.tex
minor updates, ReviewCahnges file added

@@
 
-In parallel, the OGC API Features standard \cite{ogcapifeaturespart3} is developed, which also proposes functions for filtering feature collections.
+Coincident with GeoSPARQL 1.1 development, the OGC API Features standard \cite{ogcapifeaturespart3} is being developed.
+It also proposes functions for filtering feature collections.
 While OGC API Features proposes the use of the Common Query Language (CQL) for filtering, it is also open to other query language implementations such as GeoSPARQL.
-When comparing the filter capabilities of CQL to GeoSPARQL, one can observe that concerning spatial filter capabilities, the two query languages provide comparable functionality.
-However, the version of CQL proposed by OGC API Features also provides support of spatiotemporal operators, which may be an addition to be further explored in the contiuous GeoSPARQL development process.
+When comparing the filter capabilities of CQL to GeoSPARQL, one can observe that the two query languages provide comparable spatial functionality, however the CQL proposed also supports spatiotemporal operators, which may be an addition to GeoSPARQL to be further explored in its contiuous development process.
 

### 2021-04-21 manuscript.tex
7 pages

@@
     {
-    SURROUND Australia Pty Ltd., Australia \&\\
+    %SURROUND Australia Pty Ltd., Australia \&\\
     Australian National University, Australia\\
@@
 In 2012 the Open Geospatial Consortium published GeoSPARQL 
-defining ``SPARQL extension functions'', ``RIF rules'' and ``an RDF/OWL ontology for [spatial] information''.\\
-
+defining ``SPARQL extension functions'', ``RIF rules'' and ``an RDF/OWL ontology for [spatial] information''.
 In the 8+ years since its publication, GeoSPARQL has become the most important spatial Semantic 
@@
 
-An update to the standard was proposed in 2019 to deliver GeoSPARQL 1.1 in 2021 with a charter to: 
+An update to GeoSPARQL was proposed in 2019 to deliver  1.1 in 2021 with a charter to: 
 handle outstanding change requests and source new ones from the user community and 
@@
 The GeoSPARQL standard, issued in 2012 by the Open Geospatial Consortium (OGC)\footnote{\url{https://www.ogc.org}} 
-is one of, if not the most, popular \textit{Semantic Web} standards for spatial data.\footnote{References to GeoSPARQL in other well-known standards, 
+is one of the most popular \textit{Semantic Web} standards for spatial data.\footnote{References to GeoSPARQL in other well-known standards, 
 such as DCAT2 (\url{https://www.w3.org/TR/vocab-dcat/}) and CIDOC-CRM \cite{doerr2013crmgeo} suggests it is popular, as do the incoming links in Linked Open Vocabularies \url{https://lov.linkeddata.es/dataset/lov/vocabs/gsp} and the list of implementers that includes most of the popular triplestore vendors, a list of which has been compiled here: \url{https://github.com/opengeospatial/ogc-geosparql/issues/59}}
@@
 
-In this publication we discuss the motivations behind updating GeoSPARQL 1.0 in \Cref{sec:motivation}, the content of the planned updates for the upcoming GeoSPARQL 1.1 release in \Cref{sec:newfeatures} and finally in \Cref{sec:conclusions} provide an outlook to further feature requests which are likely to be tackled in future GeoSPARQL 1.2 and 2.0 releases.
+Here we discuss the motivations behind updating GeoSPARQL 1.0 in \Cref{sec:motivation}, content of the planned GeoSPARQL 1.1 release in \Cref{sec:newfeatures} and finally in \Cref{sec:conclusions} provide an outlook to further feature requests which are likely to be tackled in future GeoSPARQL 1.2 and 2.0 releases.
 
@@
 
-The SWG's charter - its final scope of work - is also published by the OGC~\cite{abhayaratna2020ogc} and this guides 
+The SWG's charter - final scope of work - is also published by the OGC~\cite{abhayaratna2020ogc} and this guides 
 the SWG's activities. Specific actions of the SWG and their staging are explained through the use of a publicly-available 
 online task tracking system within the SWG's working online code repository\footnote{\url{https://github.com/opengeospatial/ogc-geosparql/projects/1}}.
-
 At a high-level, proposed updates to GeoSPARQL by both the SDWWG and the SWG may be categorised as one of the following:
@@
     \begin{itemize}
-        \item[$-$] GeoJSON, KML and other popular formats
+        \item GeoJSON, KML and other popular formats
     \end{itemize} 
@@
     \begin{itemize}
-        \item[$-$] implementing functions well-known in non \texttt{Semantic Web} spatial systems
+        \item implementing functions well-known in non \texttt{Semantic Web} spatial systems
     \end{itemize} 
-    \item scalar spatial properties 
-    \begin{itemize}
-        \item[$-$] area, volume etc. alongside geometries
-    \end{itemize} 
+    \item scalar spatial properties (area, volume etc. alongside geometries)
     \item better handling of Spatial (Coordinate) Reference Systems (SRS)
     \begin{itemize}
-        \item[$-$] potentially allowing for automated coordinate serialization conversions
+        \item potentially allowing for automated coordinate serialization conversions
     \end{itemize} 
@@
 points above as expected or potential.
-
 The SWG's \textit{Charter}, anticipating that the more obvious updates such as new geometry serializations would certainly
@@
 \begin{itemize}
-    \item a revision of the ``upper ontology'' structuring of GeoSPARQL
-    \begin{itemize}
-        \item[$-$] better/differently defining how GeoSPARQL's \texttt{Feature} and other classes relate to one another and to other fundamental concepts in the ontology
-    \end{itemize} 
+    \item revising ``upper ontology'' GeoSPARQL structure - how its classes relate to fundamental concepts in ontology
     \item alignments to other ontologies, perhaps \textit{W3C Time Ontology in OWL}~\cite{simon_cox_time_2017}
@@
     \item \textbf{1.2}: fully or mostly compatible extensions but which are larger additions to the standard's conceptual coverage
-    \item \textbf{2.0}: a future GeoSPARQL that might be quite different and partly incompatible with GeoSPARQL 1.0
+    \item \textbf{2.0}: future GeoSPARQL likely incompatible with GeoSPARQL 1.0
 \end{itemize}
@@
 
-An originally unforseen area of updates to GeoSPARQL is new modes of standard presentation. Motivatied by conceptual work within the W3C and the OGC for  
-multi-part standards presentation, this has resulted in profile declarations, explained 
-in the next section.
+Originally unexpected, an area of updates to standard presentation. Motivatied by conceptual work within the W3C and OGC for  
+multi-part standards, this has resulted in profile declarations explained in the next section.
 
 \section{Updates in GeoSPARQL 1.1}\label{sec:newfeatures}
-So far (as of April, 2021) the GeoSPARQL SWG has triaged change requests for GeoSPARQL releases and has addressed
-many 1.1 requests which are the only ones we report here and which are accessible through the living document for the GeoSPARQL 1.1 standard \cite{ogcgeosparql11}.
+So far (April, 2021) the GeoSPARQL SWG has triaged change requests for GeoSPARQL and has addressed
+many 1.1 requests - the only ones we report here and which are accessible through the living document for the GeoSPARQL 1.1 standard \cite{ogcgeosparql11}.
 The next sections foreshadow likely 1.2 and 2.0 updates.
@@
 \subsection{Profile Declaration}\label{sec:profiledec}
-One of the first actions undertaken by the SWG was to link the GeoSPARQL 1.0 elements through a \textit{profile} 
-declaration, where a profile is a special type of \textit{specification}, as defined by \textit{The Profiles Vocabulary}~\cite{atkinson_profiles_2020}. 
-The specific motivation for this was the SWG's recognition that GeoSPARQL 1.0 consisted of multiple parts, not all
-of which were easy to discover. As a result, some GeoSPARQL users were unaware of some of the resources and some
+One of the first SWG actions was to link GeoSPARQL 1.0 elements through a \textit{profile} 
+declaration, where a profile is a type of \textit{specification}, as defined by \textit{The Profiles Vocabulary}~\cite{atkinson_profiles_2020}. 
+Motivation for this was the SWG's recognition that GeoSPARQL 1.0 consisted of multiple parts, not all
+of which were easy to discover, so some GeoSPARQL users were unaware of some resources and some
 resources were accidentally duplicated or partly re-implemented. Profile declarations of this sort are anticipated, by the OGC, 
-as being the \textit{best practice} way for it to deliver multi-part standards.
-
-The profile declaration for GeoSPARQL 1.0 will be published by the OGC as a stand-alone resource sometime in early 2021 along with some 
-updated GeoSPARQL 1.0 resources. The profile declaration for GeoSPARQL 1.1 will be published at the same time as the 
+as being the \textit{best practice} way for multi-part standards delivery.
+The profile declaration for GeoSPARQL 1.0 will be published as a stand-alone resource along with some 
+updated GeoSPARQL 1.0 resources and at the same time as the 
 1.1 releases, currently expected in mid-2021. All 1.0 and 1.1 release resources are currently available in draft form 
@@
     \item a Simple Features feature types vocabulary
-    \item a set of RIF~\cite{kifer2013rif} rules
+    \item a set of \textit{Rules Interchange Format} rules
     \item SHACL~\cite{knublauch_shapes_2017} shapes for RDF data validation 
@@
 \begin{enumerate}
-    \item \textbf{GeoJSON} (Geo- JavaScript Object Notation)~\cite{butler2016geojson}
-    \item \textbf{KML} (Keyhole Markup Language)~\cite{nolan2014keyhole} 
+    \item \textbf{GeoJSON} (Geo-JavaScript Object Notation)\footnote{\url{https://geojson.org}}%\cite{butler2016geojson}
+    \item \textbf{KML} (Keyhole Markup Language)%\footnote{\url{http://doi.org/10.1007/978-1-4614-7900-0\_17}}%~\cite{nolan2014keyhole} 
     \item \textbf{DGGS} (Discrete Global Grid System)~\cite{sahr1998discrete}
@@
 GeoJSON \& KML have been much anticipated and were requested by the SDWWG and many users of 
-GeoSPARQL, due to those formats' popularity.
-
-The DGGS format is more forward-looking in that it is not driven by user demand but by predicted demand.
+GeoSPARQL, due to those formats' popularity. The DGGS format is more forward-looking in that it is not driven by user demand but by predicted demand.
 DGGS does not have a single, concrete format standard as the others do, nor is it ever likely to - different DGGSes will 
@@
 as well as \href{http://www.opengis.net/def/function/geosparql/boundingCircle}{\emph{geof:boundingCircle}}, \href{http://www.opengis.net/def/function/geosparql/centroid}{\emph{geof:centroid}}, \href{http://www.opengis.net/def/function/geosparql/concatLines}{\emph{geof:ConcatLines}} - concatenating a set of overlapping linestrings
-that overlap - and \emph{geosf:ConcaveHull} that can return aggregated results. Listing~\ref{lst:agfunc} gives an example 
-of one of these new functions in use.
+that overlap - and \emph{geosf:ConcaveHull} that can return aggregated results. Listing~\ref{lst:agfunc} shows one new function in use.
 
@@
 \begin{lstlisting}[caption=Aggregation Function example SPARQL query,label=lst:agfunc,language=sql,frame=single,basicstyle=\ttfamily,showstringspaces=false]
-# returns the circle geometry bounding all the geometries 
-# of Feature <x>
+# returns circle bounding all geometries of Feature <x>
 SELECT (geof:BoundingCircle(?geo) AS ?circ)
-WHERE {
-    <x> geo:hasGeometry/geo:asWKT ?geo .
-}
+WHERE {<x> geo:hasGeometry/geo:asWKT ?geo .}
 \end{lstlisting}
@@
 
-Coincident with GeoSPARQL 1.1 development, the OGC API Features standard \cite{ogcapifeaturespart3} is being developed.
-It also proposes functions for filtering feature collections.
-While OGC API Features proposes the use of the Common Query Language (CQL) for filtering, it is also open to other query language implementations such as GeoSPARQL.
-When comparing the filter capabilities of CQL to GeoSPARQL, one can observe that the two query languages provide comparable spatial functionality, however the CQL proposed also supports spatiotemporal operators, which may be an addition to GeoSPARQL to be further explored in its contiuous development process.
+Coincident with GeoSPARQL 1.1 development, the OGC API Features standard \cite{ogcapifeaturespart3} is being developed that proposes feature collections functions filtering. 
+While it proposes the use of the Common Query Language (CQL) for filtering, it is also open to other query language implementations such as GeoSPARQL. When comparing the 
+filter capabilities of CQL to GeoSPARQL, one can observe that the two query languages provide comparable spatial functionality, however the CQL proposed supports 
+spatiotemporal operators, which may be an addition to GeoSPARQL to be further explored in its contiuous development process.
 
@@
 \href{http://www.opengis.net/ont/geosparql#hasArea}{\texttt{geo:hasArea}}, \href{http://www.opengis.net/ont/geosparql#hasLength}{\texttt{geo:hasLength}} and \href{http://www.opengis.net/ont/geosparql#hasVolume}{\texttt{geo:hasVolume}} which make these attributes of a geometry better accessible using
-SPARQL.
-
-These additions address requests from the SDWWG \& SWG but
-do more too: they open up GeoSPARQL to general patterns of measurement present in ontologies 
+SPARQL. These additions address requests from the SDWWG \& SWG but
+also open up GeoSPARQL to general patterns of measurement present in ontologies 
 such as the W3C's \textit{SOSA}~\cite{haller_semantic_2017}. Similarly, 

### 2021-08-20 manuscript.tex
copyright notice

@@
 \usepackage{cleveref}
+\usepackage{tikz}
 
@@
 
+\newcommand\copyrighttext{%
+  \footnotesize Copyright ©2021 for this paper by its authors. Use permitted under Creative Commons License Attribution 4.0 International (CC BY 4.0).}
+\newcommand\copyrightnotice{%
+\begin{tikzpicture}[remember picture,overlay]
+\node[anchor=south,yshift=10pt] at (current page.south) {\fbox{\parbox{\dimexpr\textwidth-\fboxsep-\fboxrule\relax}{\copyrighttext}}};
+\end{tikzpicture}%
+}
+
 \begin{document}
 
-\title{GeoSPARQL 1.1: an almost decadal update to the most important geospatial LOD standard}
+\title{GeoSPARQL 1.1: An Almost Decadal Update to the Most Important Geospatial LOD Standard}
 \titlerunning{GeoSPARQL 1.1}
@@
 \maketitle
+
+\copyrightnotice
 
